---
title: "3. Introduction to hypothesis testing via binomial tests"
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these include showing output by default and looking for files relative to .Rproj file, not .qmd file, which makes putting files in different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Remember you should

-   add code chunks by clicking the *Insert Chunk* button on the toolbar
    or by pressing *Ctrl+Alt+I* to answer the questions!
-   use visual mode or **render** your file to produce a version that
    you can see!
-   **render** your file to make sure it runs (and that you haven't been
    working out of order)
-   save your work often
    -   **commit** it via git!
    -   **push** updates to github

## Overview

This practice reviews the [Hypothesis testing starting with binomial
tests
lecture](https://jsgosnell.github.io/cuny_biostats_book/content/chapters/Binomial.html){target="_blank}.

## Hypothesis Testing and the Binomial Distribution

### Example

Using the bat paper from class (Geipel et al. 2021), let's consider how
to analyze data showing all 10 bats chose the walking over the
motionless model.

```{r}
binom.test(10,10)
```

We use the binom.test function. We only need arguments for \# of
succeses and \# of trials. By default it runs a 2-sided test against a
null hypothesis value of p = .5. You can see how to update thee options
by looking at the help file.

```{r, eval=F}
?binom.test
```

Note the confidence interval is assymetric since its estimated to be 1!
We can see other options using the binom.confint function from the
*binom* package.

```{r}
library(binom)
binom.confint(10,10)
```

All of these correct for the fact that most intervals use a normal
approximation, which as you remember from our earlier discussions is not
good when sample sizes are small and/or the p parameter is extreme
(close to 0 or 1).

## Swirl lesson

Swirl is an R package that provides guided lessons to help you learn and
review material. These lessons should serve as a bridge between all the
code provided in the slides and background reading and the key functions
and concepts from each lesson. A full course lesson (all lessons
combined) can also be downloaded using the following instructions.

**THIS IS ONE OF THE FEW TIMES I RECOMMEND WORKING DIRECTLY IN THE
CONSOLE! THERE IS NO NEED TO DEVELOP A SCRIPT FOR THESE INTERACTIVE
SESSIONS, THOUGH YOU CAN!**

-   install the "swirl" package

-   run the following code once on the computer to install a new course

    ```{r, eval=FALSE}
    library(swirl)
    install_course_github("jsgosnell", "JSG_swirl_lessons")
    ```

-   start swirl!

    ```{r, eval=F}
    swirl()
    ```

-   then follow the on-screen prompts to select the JSG_swirl_lessons
    course and the lessons you want

    -   Here we will focus on the **Hypothesis testing starting with
        binomial tests** lesson

-   TIP: If you are seeing duplicate courses (or odd versions of each),
    you can clear all courses and then re-download the courses by

    -   exiting swirl using escape key or bye() function

        ```{r, eval=F}
        bye()
        ```

    -   uninstalling and reinstalling courses

        ```{r, eval=F}
        uninstall_all_courses()
        install_course_github("jsgosnell", "JSG_swirl_lessons")
        ```

    -   when you restart swirl with swirl(), you may need to select

        -   No. Let me start something new

## Practice!

Make sure you are comfortable with null and alternative hypotheses for
all examples.

### 1

Are people eared (do they prefer one ear or another)? Of 25 people
observed while in conversation in a nightclub, 19 turned their right ear
to the speaker and 6 turn their left ear to the speaker. How strong is
the evidence for eared-ness given this data (adapted from Analysis of
Biological Data)?

-   state a null and alternative hypothesis
-   calculate a test statistic (signal) for this data
    -   Make sure you understand how to construct a null distribution
        -   using sampling/simulation (code or written explanation)
        -   by using an appropriate distribution (code or written
            explanation)
-   Calculate and compare p-values obtained using
    -   simulation (calculation won’t be required on test, but make sure
        you understand!) (code or written explanation)
    -   equations for binomial distribution (code or written
        explanation) + R functions (required)(code)
-   Calculate a 95% confidence interval for the proportion of people who
    are right-eared
    -   How do your 95% confidence interval and hypothesis test compare?

### 2

A professor lets his dog take every multiple-choice test to see how it
compares to his students (I know someone who did this). Unfortunately,
the professor believes undergraduates in the class tricked him by
helping the dog do better on a test. It’s a 100 question test, and every
questions has 4 answer choices. For the last test, the dog picked 33
questions correctly. How likely is this to happen, and is there evidence
the students helped the dog?

**MAKE SURE TO THINK ABOUT YOUR TEST OPTIONS**
