---
title: "Compare means among groups"
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Remember you should

-   add code chunks by clicking the *Insert Chunk* button on the toolbar
    or by pressing *Ctrl+Alt+I* to answer the questions!
-   use visual mode or **render** your file to produce a version that
    you can see!
-   **render** your file to make sure it runs (and that you haven't been
    working out of order)
-   save your work often
    -   **commit** it via git!
    -   **push** updates to github

## Overview

This practice reviews the [Compare means among groups
lecture](https://jsgosnell.github.io/cuny_biostats_book/content/chapters/Compare_means_among_populations.html){target="\"_blank"}.

## Examples

We will run ANOVA's using the *lm* function to connect them to other
test. First, build the model

```{r}
iris_anova <- lm(Sepal.Length~Species, iris)
```

Then use the object it created to test assumptions

```{r}
par(mfrow = c(2,2))
plot(iris_anova)
```

If assumptions are met, check the p-value using the *summary* or *Anova*
function.

```{r}
summary(iris_anova)
library(car)
Anova(iris_anova, type = "III")
```

If the overall test is significant, carry out post hoc tests (Tukey
shown here for all pairs, as most common)

```{r}
library(multcomp)
compare_cont_tukey <- glht(iris_anova, linfct = mcp(Species = "Tukey"))
summary(compare_cont_tukey)
```

Note a special case of an ANOVA (which is a special case of a linear model) is 
when we only are comparing 2 populations. When this happens, we have a t-test. 
For example, we can compare only *I. virginica* and *I. setosa*. 

```{r}
not_versicolor <- iris[iris$Species != "versicolor",]
t.test(Sepal.Length ~ Species, not_versicolor)
```
If you compare lm and ANOVA output, it may look slightly different.

```{r}
summary(lm(Sepal.Length ~ Species, not_versicolor))
```
That's because 
by default the t-test does not assume equal variances. We can fix that.

```{r}
t.test(Sepal.Length ~ Species, not_versicolor, var.equal=T)
```

but in this case it has minimal impact.  

If assumptions are not met, we can use the Kruskal Wallis non-parametric
test and associated post hoc tests.

```{r}
kruskal.test(Sepal.Length ~ Species, data = iris)
pairwise.wilcox.test(iris$Sepal.Length, 
                          iris$Species, 
                          p.adjust.method="holm")
```

The 2-sample connected test here is a Wilcoxon test (also known as the 
Mann-Whitney test or Mann-Whitney-Wilcox test) (and why we used pairwise.wilcox.test
earrlier).

```{r}
wilcox.test(Sepal.Length ~ Species, not_versicolor, var.equal=T)
```

We can also use a bootstrap alternative

```{r}
library(WRS2)
t1waybt(Sepal.Length~Species, iris)
bootstrap_post_hoc <- mcppb20(Sepal.Length~Species, iris)
p.adjust(as.numeric(bootstrap_post_hoc$comp[,6]), "holm")
```

For 2 groups, the *boot.t.test* function in the **MKinfer** package is
also an option.

```{r}
library(MKinfer)
boot.t.test(Sepal.Length ~ Species, not_versicolor)
```

A final option (shown here for 2 groups) is to use permutation tests.  These
tests work by rearranging the data in all (theoretically) possible ways and 
calculating signals for each permutation.  You can then compare what we actually 
found to that to obtain p-values.  Often times we can't find all the combinations 
there are too many!), but a good sample (10,000+) should work. The 
function that we'll use for this is independence_test from the coin package. 
It uses the same arguments as the t-test (including the formula interface). 

```{r}
library(coin)
independence_test(Sepal.Length ~ Species, not_versicolor)
```

## Swirl lesson

Swirl is an R package that provides guided lessons to help you learn and
review material. These lessons should serve as a bridge between all the
code provided in the slides and background reading and the key functions
and concepts from each lesson. A full course lesson (all lessons
combined) can also be downloaded using the following instructions.

**THIS IS ONE OF THE FEW TIMES I RECOMMEND WORKING DIRECTLY IN THE
CONSOLE! THERE IS NO NEED TO DEVELOP A SCRIPT FOR THESE INTERACTIVE
SESSIONS, THOUGH YOU CAN!**

-   install the "swirl" package

-   run the following code once on the computer to install a new course

    ```{r, eval=FALSE}
    library(swirl)
    install_course_github("jsgosnell", "JSG_swirl_lessons")
    ```

-   start swirl!

    ```{r, eval=F}
    swirl()
    ```

    -   swirl()

-   then follow the on-screen prompts to select the JSG_swirl_lessons
    course and the lessons you want

    -   Here we will focus on the **Compare means among groups** lesson

-   TIP: If you are seeing duplicate courses (or odd versions of each),
    you can clear all courses and then re-download the courses by

    -   exiting swirl using escape key or bye() function

        ```{r, eval=F}
        bye()
        ```

    -   uninstalling and reinstalling courses

        ```{r, eval=F}
        uninstall_all_courses()
        install_course_github("jsgosnell", "JSG_swirl_lessons")
        ```

    -   when you restart swirl with swirl(), you may need to select

        -   No. Let me start something new

## Just for practice

*For the following questions, you will use multiple methods to analyze a 
single dataset. This is for practice and so you can compare! For actual 
analysis you should only use the best method!*

### 1

Use the iris dataset in R to determine if petal length differs among
species. *Do this problem using the following methods*

-   *ANOVA*

-   *Kruskal-Wallis*

-   *bootstrapping*

*Make sure you can plot the data and carry out multiple comparison
methods as needed. Also be sure to understand the use of coefficients
and adjusted R^2^ values and where to find them.*

### 2

Data on plant heights (in cm) for plants grown with a new and old
formulation of fertilizer can be found at

<https://docs.google.com/spreadsheets/d/e/2PACX-1vSUVowOKlmTic4ekL7LSbwDcqrsDSXv5K_c4Qyfcvz1lLE1_iINmGzy0zMGxY7z5DImlUErK4S2wY7Y/pub?gid=0&single=true&output=csv>.

Analyze this data using the t.test function and the lm function to
convince yourself that t-tests are special cases of ANOVAs, which are
special cases of linear models!

*For the following questions, pick the appropriate method for analyzing the question. Use a plot of the data and/or model analysis to justify your decision. Make sure you can carry out multiple comparison methods as needed. Also be sure to understand the use of coefficients and adjusted R^2^ values and where to find them.*

For the following questions, pick the appropriate method for analyzing the question. Use a plot of the data and/or model analysis to justify your decision. Make sure you can carry out multiple comparison methods as needed. Also be sure to understand the use of coefficients and adjusted R2 values and where to find them.

### 3

Data on sugar cane yield for multiple fields is available using

read.table("<https://docs.google.com/spreadsheets/d/e/2PACX-1vRjstKreIM6UknyKFQCtw2_Q6itY9iOAVWO1hUNZkBFL8mwVssvTevqgzV22YDKCUeJq0HBDrsBrf5O/pub?gid=971470377&single=true&output=tsv>",
header = T, stringsAsFactors = T)

More info on the data can be found at
<http://www.statsci.org/data/oz/cane.html>. Is there evidence that
location (DistrictPosition column) impacts yield (Tonn.Hect column)? If
so, which areas are driving this distance?

### 4

Data on FEV (forced expiratory volume), a measure of lung function, can
be found at

<http://www.statsci.org/data/general/fev.txt>

More information on the dataset is available at

<http://www.statsci.org/data/general/fev.html>.

Is there evidence that FEV depends on gender? If so, which gender has
the higher FEV score? How much variance does gender explain?

### 5

The following data are human blood clotting times (in minutes) of
individuals given one of two different drugs.

| Drug B | Drug G |
|:------:|:------:|
|  8.8   |  9.9   |
|  8.4   |  9.0   |
|  7.9   |  11.1  |
|  8.7   |  9.6   |
|  9.1   |  8.7   |
|  9.6   |  10.4  |
|        |  9.5   |

Test the hypothesis that the mean clotting times are equal for the two
groups

-   Estimating the variance from the data
-   Using rank transform analysis
-   Using a permutation test
-   Using a bootstrap test

Test the hypothesis that the mean clotting times are equal for the two
groups

-   Estimating the variance from the data
-   Using rank transform analysis
-   Using a permutation test
-   Using a bootstrap test

### 6

(Example from Handbook on Biological Statistics) Odd (stunted, short,
new) feathers were compared in color to typical feathers in Northern
Flickers (*Colaptes auratus*) [@wiebe2002] . Data is at

<https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/wiebe_2002_example.csv>

Test the hypothesis that odd and typical feathers did not differ using

-   a Student's t test and/or lm
-   a rank test
-   bootstrapping

Note we will return to this question next week!

### 7

Find an example of a one-way ANOVA from a paper that is related to
your research or a field of interest (hint: use Google Scholar or another search engine to search for a keyword of interest plus ANOVA).  Make sure you understand the connections between the methods,
results, and graphs. Briefly answer the following questions

-   What was the dependent variable?
-   What were the independent variables?
-   Was there a difference in mean values among the groups? If so, how did the paper address or explain this finding.
