---
title: "Relationships among numerical variables"
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Remember you should

-   add code chunks by clicking the *Insert Chunk* button on the toolbar
    or by pressing *Ctrl+Alt+I* to answer the questions!
-   **render** your file to produce a markdown version that you can see!
-   save your work often
    -   **commit** it via git!
    -   **push** updates to github

## Overview

This practice reviews the [Relationships among numerical variabless
lecuture](https://jsgosnell.github.io/cuny_biostats_book/content/chapters/Relationships_among_numerical_variables.html){target="_blank"}.

## Example

Following the iris dataset from class, let's investigate the
relationships between sepal and petal length. First, we can plot the
data.

```{r}
library(ggplot2)
ggplot(iris, aes(x=Petal.Length, y=Sepal.Length)) +
  geom_point(size = 3) +
  ylab("Sepal Length")+ggtitle("Sepal length increases with petal length")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+
  xlab("Petal length (cm)") +
  ylab("Sepal length (cm)")

```

A linear relationship appears logical here. We can build the model and
check assumptions.

```{r}
iris_regression <- lm(Sepal.Length ~ Petal.Length, iris)
plot(iris_regression)

```

Visual checks of assumptions appear to be met, so we can determine if
the slope differs from 0.

```{r}
library(car)
Anova(iris_regression, type = "III")
summary(iris_regression)
```

We find that F~1,148~=468.5, p \< 0.01, so we reject the null hypothesis
that the slope is equal to 0. The estimate indicates a slope of 0.408,
so sepal length increases with petal length. An R^2^ value of 0.76
indicates petal length explains about 75% of the variation in sepal
length.

Alternatively, we could consider if the association between the two
variables is equal to 0 (or not).

```{r}
cor.test(~ Sepal.Length + Petal.Length, data = iris)

```

We find the same p-value (using a t-distribution), and see that the
estimated linear correlation coefficient, .87, is the square of our R^2^
value.

If we prefer a rank-based test, we can update the code:

```{r}
cor.test(~ Sepal.Length + Petal.Length, data = iris,
         method="spearman")
```

Which again leads us to reject the null hypothesis. Finally, a bootstrap
options can be produced

```{r}
bootstrap_iris <- Boot(iris_regression)
Confint(bootstrap_iris)
```

where estimates indicated the slope does not contain 0.

## Practice

### 1

A professor carried out a long-term study to see how various factors
impacted pulse rate before and after exercise. Data can be found at

http://www.statsci.org/data/oz/ms212.txt

With more info at

http://www.statsci.org/data/oz/ms212.html.

Is there evidence that age, height, or weight impact change in pulse
rate for students who ran (Ran column = 1)? For each of these, how much
variation in pulse rate do they explain?

### 2

(from OZDASL repository,
http://www.statsci.org/data/general/stature.html; reference for more
information)

When anthropologists analyze human skeletal remains, an important piece
of information is living stature. Since skeletons are commonly based on
statistical methods that utilize measurements on small bones. The
following data was presented in a paper in the American Journal of
Physical Anthropology to validate one such method. Data is available \@

http://www.statsci.org/data/general/stature.txt

as a tab-delimted file (need to use read.table!) Is there evidence that
metacarpal bone length is a good predictor of stature? If so, how much
variation does it account for in the response variable?

### 3

Data on medals won by various countries in the 1992 and 1994 Olympics is
available in a tab-delimited file at

http://www.statsci.org/data/oz/medals.txt

More information on the data can be found at:

http://www.statsci.org/data/oz/medals.html

Is there any relationship between a country's population and the total
number of medals they win?

### 4

Continuing with the Olympic data, is there a relationship between the
latitude of a country and the number of medals won in summer or winter
Olympics?

### 5

Data on FEV (forced expiratory volume), a measure of lung function, can
be found at

<http://www.statsci.org/data/general/fev.txt>

More information on the dataset is available at

<http://www.statsci.org/data/general/fev.html.>

Is there evidence that FEV depends on age or height? If so, how do these
factors impact FEV, and how much variance does each explain?

### 6

Continuing with the FEV data, produce plots that illustrate how height,
age, and gender each impact FEV.

### 7

Find an example of a regression or correlation from a paper that is related to
your research or a field of interest.  Make sure you understand the connections between the methods,
results, and graphs. Briefly answer the following questions

-   What was the dependent variable?
-   What were the independent variables?
-   Was there a relationship between them? If so, describe it (positive, negative).
