---
title: Swirl lessons
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

[Swirl](https://swirlstats.com/){target="_blank"} is a program/package
that can help teach you R at your own pace. There are many swirl lessons
that are freely available (most of which I have not tried!). Given that,
you are free to explore them, but I have developed a set to follow our
class (with support of a [QUBES working
group](https://qubeshub.org/community/groups/teaching_r_fmn){target="_blank"}).
To work with them, you'll need to

-   install swirl

    -   You can use the Install Packages option in Rstudio or type the
        following into your console (or copy and paste it)!

```{r}
install.packages("swirl")
```

-   load the package and install the lessons for the book
    -   This will install swirl, load the package, and load the lessons.

```{r}
library(swirl)
install_course_github("jsgosnell", "JSG_swirl_lessons")
```

-   start swirl

```{r}
swirl()
```

-   Follow the on-screen prompts to select the CUNY_swirl course and the
    lesson you want to focus on. Lesson names match the chapter names.
    Note not all chapters have a matching swirl lesson!

Occasionally I update the lessons. If you want the latest version,
you'll need to uninstall the course

```{r}
uninstall_course("JSG_swirl_lessons")
```

and then reinstall it following the original instructions.

