---
title: One sample tests for continuous data
subtitle: The one where we fully introduce normality
bibliography: references.bib
---

In this chapter will build on our introduction to significance testing
by considering tests for continuous data collected on one trait from a
single population. This will also allow/require us to more fully define
*normal* distributions, which we have already started to discuss.

## Example

Let's return to our iris data and focus on sepal lengths of *I.
viriginica*.

```{r}
library(ggplot2)
ggplot(iris[iris$Species == "virginica",],
              aes(x=Sepal.Length)) +
  geom_histogram( fill="blue", color="black") +
  labs(title=expression(paste("Sepal lengths of ",italic("I. virginica"))),
       x= "Sepal length (cm)",
       y= "Frequency")
```

What if we wanted to test if the height was equal to a certain value
like 7 cm?

**We can't, and it's not**.

Height is a random variable. It differs among individuals (see above),
so it isn't equal to any specific value. This may seem obvious, but it's
an important step in understanding hypothesis testing. Many students
also struggle with this, but most are with the fact we learned about
hypothesis testing focusing on proportions (remember the last chapter?).
When we focused on binomial data, it was obvious a single draw could not
be 2 things - it was a success or failure, and we focused on the
relative occurrence of those.

Similarly, for continuous numeric data (remember: data that can on any
value in a given range), we need to focus on describing the distribution
of the data. If we do that, we might want (and be able to test) if, for
example, the *mean* height of *I. virginica* is equal to 7 cm. In fact,
we typically focus on the *mean* of the distribution (one of measures of
central tendency)

To do this, we need to do what we did with binomial data: develop a null
hypotheses, use it to construct a null distribution, and compare our
data to it see how unusual it is (and get a p-value).

<details>

<summary>What is our null hypothesis for this example?</summary>

For this example, we are focused on a two-tailed test (we are asking if
the mean is equal to a certain value), so we have

$$
\begin{split}
H_O: \mu_{height} = 7 \ cm \\ 
H_A: \mu_{height} \neq 7 \ cm
\end(split)
$$

</details>

Now that we have a null hypothesis, we need to test it. We can do this
by simulation. Let's make a distribution where the
$\mu_{height} =7 \  cm$, then draw samples from it and see how rare it
is to get what we actually observed in the data...which was

```{r}
mean(iris[iris$Species == "virginica","Sepal.Length"])
```

Seems easy enough, but what distribution do we draw from? For our
binomial data we knew exactly what to parameterize - that's because the
entire distribution is described by the parameter *p* (go back to the
last chapter and note we can find the spread using this one variable as
well!).
