---
title: Maps in R
subtitle: Where we meet the p-value!
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

Making maps is something many scientists need to do. Understanding the spatial
relationship among sampling sites, populations, or cities can inform our understanding
of analysis or results.  While R is not primarily a geographic information
system, it has the ability to produce maps through multiple libraries (the focus
of this section). It can also be used to perform spatial analysis.

## Making maps

Producing maps typically involves loading a background/base map of an area and then
plotting sites or specific data points on it.  For this to work, you have to make
sure the different maps are using the same coordinate system and map projection.
Put simply, coordinates (e.g., latitude and longitude) are used to specify a spot 
on the earth. Various coordinate systems exist, so you need to make sure the base
map is using the same coordinates you are using to specify location.  Part of 
different coordinate systems is related to the fact that dislaying a spherical 
earth on a flat screen/piece of paper is hard.  
In order for it to end a "complete picture", most projections distort various
pieces of th earth. 

Map data can be produced using vector or raster data. Vectors mean the software is
using directions to draw points, lines, or polygons that represent space.  Raster
means the software is actually stitching together cells or images to represent space.

### Vector approaches

For example, rnaturalearth supplies vector data from NaturalEarth for use in R.
You can ask for country level data. Note that the data contains info on how to
"draw" each 
country, but can also hold other data.

```{r}
library(rnaturalearth)
base_map <- ne_countries(type = 'countries', scale = 'small')
library(rmarkdown)
paged_table(base_map@data)
```

this can be plotted immediately

```{r}
plot(base_map)
```

Other commands only focus on coastlines

```{r}
plot(ne_coastline())
```

You can also turn this object into a dataframe (fortify) and use it in ggplot2

```{r}
library(ggplot2)
base_map_fortified=fortify(base_map)
names(base_map_fortified)[names(base_map_fortified) == "long"] <- "Longitude"
names(base_map_fortified)[names(base_map_fortified) == "lat"] <- "Latitude"

ggplot(data=base_map_fortified,aes(x=Longitude, y=Latitude, group=group)) +
  geom_polygon(colour="black", fill="white") +
  xlab("Longitude")+
  ylab("Latitude") 

```

Once you have these layers, you can start adding your own sites

```{r}
sites <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/SurveySiteCoordinates.csv", strip.white = T, stringsAsFactors = T)
ggplot(data=base_map_fortified,aes(x=Longitude, y=Latitude, group=group)) +
  geom_polygon(colour="black", fill="white") +
  xlab("Longitude")+
  ylab("Latitude") +
  geom_point(data = sites, 
             aes_string(x="Longitude",y="Latitude", group = NA), size = 10)
```


### Raster approaches

Note using google maps now requires an api.  This can be created @ , but should
not be shared with others.  To use yours here, save it in a file called

```{r}
source("content/extensions/private_info.R")
```

```{r}

map <- get_map(location =c(lon = mean(sites$Longitude), lat = mean(sites$Latitude)),
               zoom = 12, maptype = "satellite")
#toner is another option
ggmap(map)+ 
  # xlab(expression(paste("Longitude ", ( degree~W)))) +
  # ylab(expression(paste("Latitude ", ( degree~N)))) +
  xlab(expression(paste("Longitude"))) +
  ylab(expression(paste("Latitude"))) +
  geom_point(data = bysite, 
             aes(x=Longitude, y=Latitude, group = NA, 
                 color = Location), 
             shape = 21, size = 5, stroke = 1.5) +
  #scale_x_continous (labels=abs) to get rid of negatives
  # scale_x_continuous(labels=abs) +
  scale_color_manual(values = c("white", "yellow")) +
  annotate("text", label="Yellow Bar Hassock", x=-73.85, y=40.59, size=10, 
           #          color =  hue_pal()(2)[2]) +
           color = "yellow") +
  annotate("text", label="Black Bar Marsh", x=-73.82, y=40.64, size=10, 
           #           color =  hue_pal()(2)[1]),
           color = "white") 
```


## Next steps

R can also be  used to interact with more map-focused software like QGIS.



