---
title: "4. Continuous tests for 1 population"
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* use visual mode or **render** your file to produce a version that you can see!
* **render** your file to make sure it runs (and that you haven't been working
out of order)
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Overview

This practice reviews the [Tests for continuous data from one sample
lecture](../chapters/Tests_for_continuous_data_from_one_sample.qmd){target="_blank"}.

## Examples

From lecture! Consider if average height of males training at the Australian 
Institute of Sport is different than average of human population. 

These are all one sample tests, but they differ in what we know. If we know the
variance of our population, we use a z test (function in BSDA package).

```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
library(BSDA)
z.test(sport[sport$Sex == "male", "Ht"], mu = 175.6, sigma.x=7)
```
If we don't, we use a t-test

```{r}
t.test(sport[sport$Sex == "male", "Ht"], mu = 175.6)
```

These both assume the means of the data are normal! If we want to relax that 
assumption, we can use the Wilcoxon test (also known as Mann-Whitney test, signed
binary transform, or other terms!). This assumes the distribution of means is
symmetric.

```{r}
wilcox.test(sport[sport$Sex == "male", "Ht"], mu = 175.6)
```

or the sign-test/media test.

```{r}
SIGN.test(sport[sport$Sex == "male", "Ht"], md = 175.6)
```

Note this is just transforming data to 1/0 and doing a binomial test!

```{r}
above_175.6 <- nrow(sport[sport$Sex == "male" & sport$Ht > 175.6,])
binom.test(above_175.6, nrow(sport[sport$Sex == "male",]))
```


We can also bootstrap the data.

```{r}
number_of_simulations <- 1000
library(ggplot2)
boostrap_data<- sport[sport$Sex == "male", "Ht"]
boostrap_outcomes <- data.frame(mean = rep(NA, number_of_simulations), sd = NA)
for (i in 1:number_of_simulations){
iris_bootstrap <-sample(boostrap_data, length(boostrap_data), replace = T)
boostrap_outcomes$mean[i] <- mean(iris_bootstrap)
boostrap_outcomes$sd[i] <- sd(iris_bootstrap)
}
ggplot(boostrap_outcomes, aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title=expression(paste("Bootstrapped means")),
       x= "Mean value",
       y= "Frequency")
```

and find associated quantile-based 95% confidence intervals:
```{r}
quantile(boostrap_outcomes$mean, probs=c(.025, .975) ) 
```

or using functions in the *boot* library
```{r}
library(boot)
results <- boot(data=boostrap_data, statistic = function(x, inds) mean(x[inds]),
   R=number_of_simulations)
ggplot(data.frame(results$t), aes(x=results.t)) +
  geom_histogram(color="black") +
  labs(title=expression(paste("Bootstrapped means")),
       x= "Mean value",
       y= "Frequency")
quantile( results$t, probs=c(.025, .975) ) 
boot.ci(results)       
```

Another option (recommended) is the *boot.t.test* function in the **MKinfer** package.

```{r}
library(MKinfer)
boot.t.test(sport[sport$Sex == "male", "Ht"], mu = 175.6)
```




## Let's practice!

### Recognizing and assessing normality 

#### 1

Using the qqplot_example.R code, examine the following distributions and, for 
the continuous distributions (marked with a “*”), observe how a normal probability
plot (qqplot) can be used to visually test for approximate normality.

* *Normal (u= 0; σ^2^= 1, 10, 100) 
* *Student's t (df = 1, 10, 30, & 100) 
* *Chi-square (df= 1, 2, 5, 30, 50)
* Bernoulli (P=0.1, 0.5, & 0.9) 
* Binomial (P=0.05; N= 2, 5, 25, & 50); (P=0.25; N= 2, 5, 25, & 50); (P=0.50; 
N= 2, 5, 25, & 50); (P=0.75; N= 2, 5, 25, & 50); (P=0.95; N= 2, 5, 25, & 50) 
* Poisson ( u= 2, 5, 10, 30, & 50)

For this question, its easiest to just source the main file and see what happens.
When you source a script, it is run in R without showing any console output
(but graphs and objects are still produced!).  Try *source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R")*

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R")
```

*Notice the spread of DATA of every distribution tend towards normality as sample size increases*

#### 2


Review the central_limit_theorem.R code if you need to convince/remind 
yourself how common normality of means is for even non-normal data. You can source the code
using the same approach noted in Question 1.

*Here we are focused on how the means look as sample size increases*

```{r}
#make sure you have VGAM library installed
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/central_limit_theorem.R")
```


### Working with data (note some sample sizes may be too small for these to all be good ideas!)

Make sure you are comfortable with null and alternative hypotheses for all examples. 
You should also feel comfortable graphing the data.

#### 3


Seven observers were shown, for a brief period, a grill with 161 flies impaled 
and were asked to estimate the number. The results are given by Cochran (1954). 
Based on five estimates, they were 183.2, 149.0, 154.0, 167.2, 187.2, 158.0, and 143.0. Test the null hypothesis that the mean of the estimates is 161 flies.

* Assuming variance = 275


```{r}
flies <- c(183.2, 149.0, 154.0, 167.2, 187.2, 158.0, 143.0)
library(BSDA)
z.test(x=flies, mu = 161, sigma.x=sqrt(275))
```
*Using a z-test, I found a test statistics of z~=0.33 .This corresponds to a p-value of 0.73. This p value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

*  Estimating the variance from the data
```{r}
t.test(x=flies,mu = 161)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal 
distribution,  I found a test statistics of t~6~=0.32.  This corresponds to a 
p-value of 0.76. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

* Using rank transform analysis 

```{r}
wilcox.test(flies, mu=161)
```
*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test
statistics of V = 15 .This corresponds to a p-value of 0.94. This p-value is >.05, so I 
fail to reject the null hypothesis that the mean of the estimates is 161 flies.*

* Using binary transform analysis
```{r}
SIGN.test(flies, md=161)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met, I found a test statistics of s = 3 .This corresponds to a p-value of 1. This p-value is >.05, so I fail to reject 
the null hypothesis that the median (**Note change here**) of the estimates is
161 flies.*

Note there are several ways to load the data!  You can make a list (since the 
list is short):

```{r}
flies <- c(183.2, 149.0, 154.0, 167.2, 187.2, 158.0, 143.0 )
```

or make a dataframe in a spreadsheet software (eg, Excel, Google Sheets) and then upload using a read.csv command. We did this in the introduction to R!

#### 4


Yields of 10 strawberry plants in a uniformity trial are given by Baker and 
Baker (1953) as 239, 176, 235, 217, 234, 216, 318, 190, 181, and 225 g. 
Test the hypothesis that µ = 205
* Assuming variance = 1500
```{r}
strawberries <- c(239, 176, 235, 217, 234, 216, 318, 190, 181, 225)
z.test(x=strawberries,mu = 205, sigma.x=sqrt(1500))
```
*Using a z-test, I found a test statistics of z=1.48.  This corresponds to a 
p-value of 0.14. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

*  Estimating the variance from the data

```{r}
t.test(x=strawberries,mu = 205)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal distribution, I found a test statistics of t~9~=1.42.  This corresponds to a 
p-value of 0.19. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using rank transform analysis 
```{r}
wilcox.test(strawberries, mu=205)
```

*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test statistics of V=40.5.  This corresponds to a 
p-value of 0.20.  This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using binary transform analysis
```{r}
SIGN.test(strawberries, md=205)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met,  I found a test statistics of s= 7.  This corresponds 
to a p-value of 0.34. This p-value is >.05,so I fail to reject 
the null hypothesis that the population median (**Note change here**) is equal to 205.*

#### 5


Evolutionary geneticists predicts the family sex ratio will be 80% female in 
broods of eagles that successfully fledge >3 young. Nests that fledge 3 or more
chicks are very rare but a sample of 30 chicks are obtained from such nests and
they yield 25 females and 5 males. Test the hypotheses that that: 
* a) the sex ratio is 50% females

```{r}
#a
binom.test(25,30, p=.5)
```
*A binomial test was used as we are comparing an observed proportion against a 
set value.  Given a p-value of <.001, I reject the null hypothesis that the 
proportion of sons is equal to .5.*

* b) the sex ratio is 80% females.

```{r}
binom.test(25,30, .8)
```
*A binomial test was used as we are comparing an observed proportion against a 
set value.  Given a p-value of <.001, I fail to reject the null hypothesis that the 
proportion of sons is equal to .8.*

#### 6


Studies of flying snakes have led researchers to posit the mean undulation 
rate is 1.4 Hz.  You wish to test this hypothesis using the small sample of 
undulation rates shown below.  Create a small dataset of the paradise tree snake 
undulation rates and choose and justify a test you can use to assess the data.

Undulation rates (in Hz):  0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6

```{r}
snakes <- c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
t.test(snakes, mu=1.4)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal
distribution,  I found a test statistics of t~7~=-.22.  This corresponds to a 
p-value of 0.83. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean undulation rate is 1.4 Hz.*

#### 7

Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in weight from the average Australian male (85.9 kg) using
bootstrapping techniques. Data at 

```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T, 
                    stringsAsFactors = T)
```


```{r}
library(MKinfer)
boot.t.test(sport[sport$Sex == "male", "Wt"], mu= 85.9)
```

*Using a bootstrap test wtih 10,000 samples, we found a p-value of .007; we thus 
reject the null hypothesis that males training at the AIS have the same weight as
the average Australian male.  Data indicated they weigh less.*

