---
title: "More ANOVAs"
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* use visual mode or **render** your file to produce a version that you can see!
* **render** your file to make sure it runs (and that you haven't been working
out of order)
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Overview

This practice reviews the [More ANOVAs lecuture](../chapters/More_ANOVAs.qmd).

## Examples

### If interaction is significant

Following the memory example from class, read in and check data

```{r}
memory <- read.table("http://www.statsci.org/data/general/eysenck.txt", header = T,
                     stringsAsFactors = T)
str(memory)
```

Let's put younger level first

```{r}
library(plyr)
memory$Age <- relevel(memory$Age, "Younger")
```

and graph

```{r}
library(Rmisc)
function_output <- summarySE(memory, measurevar="Words", groupvars =
                               c("Age", "Process"), na.rm = T)
library(ggplot2)
ggplot(function_output, aes(x=Age, y=Words,color=Process, 
                                   shape = Process)) +
  geom_line(aes(group=Process, linetype = Process), size=2) +
    geom_point(size = 5) +
  ylab("Words remembered")+ 
  xlab("Age") + 
  ggtitle("Process type interacts with age to impact memory")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

There appears to be some interactions. Let' build a model

```{r}
memory_interactions <- lm(Words ~ Age * Process, memory)
```

and check assumptions.

```{r}
par(mfrow=c(2,2))
plot(memory_interactions)
```

These appear to be met, so look at output

```{r}
library(car)
Anova(memory_interactions, type = "III")
```

Since interaction is significant, analyze subsets. For example,

```{r}
memory_interactions_young <- lm(Words ~ Process, memory[memory$Age == "Younger",])
plot(memory_interactions_young)
Anova(memory_interactions_young, type = "III")
```

There is a significant difference in words recalled based on process,
but why? Investigate with post-hoc tests.

```{r}
library(multcomp)
comp_young <- glht(memory_interactions_young, linfct = mcp(Process = "Tukey"))
summary(comp_young)
```

### Blocking example

Following feather color example from class:

```{r}
# more than 2? ####
feather <-  read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/wiebe_2002_example.csv", stringsAsFactors = T)
str(feather)
set.seed(25)
special <- data.frame(Bird = LETTERS[1:16], Feather = "Special", 
                      Color_index= feather[feather$Feather == "Typical", "Color_index"] +
                        .3 +runif(16,1,1)*.01)
feather <- merge(feather, special, all = T)


Anova(lm(Color_index ~ Feather + Bird, data=feather), type= "III")

library(multcomp)
compare <- glht(lm(Color_index ~ Feather + Bird, data=feather), linfct = mcp("Feather" = "Tukey"))
summary(compare)
```

```{r, error=T}
#note comparison doesn't work
Anova(lm(Color_index ~ Feather * Bird, data=feather), type= "III")
```

## Practice

### 1

A survey was conducted to see if athletes and non-athletes deal with
anger in the same way. Data is \@

angry \<-
read.csv("<https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv>",
stringsAsFactors = T)

and more information is at

<http://onlinestatbook.com/case_studies/angry_moods.html>.

Focus on the following variables:

Sports 1 = athletes, 2 = non-athletes Gender 1 = males, 2 = females
Expression (AE) index of general anger expression: (Anger-Out) +
(Anger-In) - (Control-Out) - (Control-In) + 48

Is there any evidence that gender or athlete status impact how anger is
expressed?

```{r}
angry <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv", stringsAsFactors = T)
str(angry)
angry$Gender <- as.factor(angry$Gender)
library(plyr)
angry$Gender <- revalue(angry$Gender, c("1" = "male", 
                                        "2" = "female"))
angry$Sports <- as.factor(angry$Sports)
angry$Sports <- revalue(angry$Sports, c("1" = "athlete",
                                        "2" = "non-athlete"))
library(Rmisc)
anger_summary <- summarySE(angry, measurevar="Anger_Expression", groupvars =
                               c("Sports", "Gender"), na.rm = T)
library(ggplot2)
ggplot(anger_summary, aes(x=Gender, y=Anger_Expression, color=Sports, 
                                   shape = Sports)) +
  geom_point(size = 3) +
  geom_line(aes(group=Sports, linetype =Sports), size=2) +
  geom_errorbar(aes(ymin=Anger_Expression-ci, ymax=Anger_Expression+ci), size=1.5) +
  ylab("Anger level")+ 
  xlab("Gender") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Anger level among groups")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
*I first read in and recoded some data for ease and plotting. I then produced a 
plot to consider the null hypotheses that* 

* *the sport an athlete plays does not influence anger level*
* *the gender of an athlete does not influence anger level*
* *the sport an athlete plays and their gender do not interact to influence anger 
level*

```{r}
angry_gender <- lm(Anger_Expression ~ Sports * Gender, angry)
plot(angry_gender)
library(car)
Anova(angry_gender, type = "III")

#remove interaction since not significant
angry_gender <- lm(Anger_Expression ~ Sports + Gender, angry)
plot(angry_gender)
Anova(angry_gender, type = "III") #only differs among those who play sports
```
*I then analyzed the data using factorial ANOVA.
The outcome is continuous 
and both explanatory variables are categorical. The design is also fully randomized. Resildual plots indicated all assumptions were met
(there is no pattern in the residuals and they are normally distributed).
Analysis shows an insignificant interaction (F~1,74~=.04, p=.855) between 
sport and gender, so I removed the interaction term. The reduced model
showed anger levels differed among athletes and non-athletes but not by 
gender. There was no need for post-hoc tests (only 2 levels/groups for
each categorial variable.)*

### 2

A professor carried out a long-term study to see how various factors
impacted pulse rate before and after exercise. Data can be found at
<http://www.statsci.org/data/oz/ms212.txt> With more info at
<http://www.statsci.org/data/oz/ms212.html>. Is there evidence that
frequency of exercise (Exercise column) and gender impact change in
pulse rate for students who ran (Ran column = 1)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, 
                    stringsAsFactors = T)
pulse$Exercise <- factor(pulse$Exercise)
library(plyr)
pulse$Exercise <- revalue(pulse$Exercise, c("1" = "high", 
                                            "2" = "moderate", 
                                            "3" = "low"))
pulse$Gender <- factor(pulse$Gender)
pulse$Gender <- revalue (pulse$Gender, c("1" = "male", "2" = "female"))
pulse$change <- pulse$Pulse2 - pulse$Pulse1
change_summary <- summarySE(pulse[pulse$Ran == 1, ], measurevar="change", groupvars =
                               c("Exercise", "Gender"), na.rm = T)

ggplot(change_summary, aes(x=Gender, shape = Exercise, color = Exercise,
                           y=change)) +
  geom_point(size = 3) +
  geom_line(aes(group=Exercise, linetype =Exercise), size=2) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Change in pulse \n (beats per minute)") +
  scale_color_discrete(name = "Exercise level")+
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Change in pulse does \n not differ among groups") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
*I first read in and recoded some data for ease and plotting. I then produced a 
plot to consider the null hypotheses that* 

* *exercise level does not influence change in pulse rate*
* *gender does not influence change in pulse rate*
* *gender and exercise level do not interact to influence change in pulse rate*

```{r}
exercise <- lm(change ~ Gender * Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III")

#rerun without interaction
exercise <- lm(change ~ Gender + Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III") #no significance
```

*I then analyzed the data using factorial ANOVA.
The outcome is continuous 
and both explanatory variables are categorical. The design is also fully randomized. Residual plots indicated all assumptions were met
(there is no pattern in the residuals and they are normally distributed).
Analysis shows an insignificant interaction (F~2,40~=1.2, p=.21) between 
exercise level and gender, so I removed the interaction term. The reduced model
showed neither gender (F~1,42~=.16, p =.69) or exercise level (F~2,42~=.1, p=.90) 
influenced change in pulse rate, so I failed to reject the related null 
hypotheses.*

### 3

Data from Valdez et al 2023 is available \@
<https://docs.google.com/spreadsheets/d/e/2PACX-1vT2gaLu6pyRMlcbzarn3ej4bFmT_iHvrlNWJYSdrsLdUWIjcJi7rU11-ipvYpGnqD9qLDnbhNd2sDUW/pub?gid=1707080634&single=true&output=csv>.

Import it into to R and

-   determine how the snail grazing and nitrogen levels impact number of
    flowering shoots(Shoot.density..m2)
-   construct a plot to showcase your analysis


```{r}
valdez_2023 <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vT2gaLu6pyRMlcbzarn3ej4bFmT_iHvrlNWJYSdrsLdUWIjcJi7rU11-ipvYpGnqD9qLDnbhNd2sDUW/pub?gid=1707080634&single=true&output=csv", stringsAsFactors = T)
shoot_model <-lm( Shoot.density..m2~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",])
plot(shoot_model)
```

```{r}
Anova(shoot_model, type="III")
```

No interaction, so I can remove or not.

```{r}
shoot_model_reduced <- update(shoot_model, .~.- Snail.Level:Nitrogen.level)
Anova(shoot_model_reduced)
```
Both have impacts.

```{r}
snail_post_hoc <- glht(shoot_model_reduced, linfct = mcp(Snail.Level = "Tukey"))
summary(snail_post_hoc)
```

significant difference among all snails.

```{r}
shoot_summary <- summarySE(valdez_2023[valdez_2023$Snail.Level != "uncaged",], measurevar = "Shoot.density..m2", groupvars = c("Snail.Level", "Nitrogen.level"))
shoot_summary
shoot_summary$Snail.Level <- relevel(shoot_summary$Snail.Level, "removal")
shoot_summary$Snail.Level <- relevel(shoot_summary$Snail.Level, "uncaged")


ggplot(shoot_summary, aes(x=Snail.Level, 
                           y=Shoot.density..m2,
                           fill=Nitrogen.level)) +
  geom_col(color="black", position=position_dodge()) +
  geom_errorbar(aes(ymin=Shoot.density..m2, ymax=Shoot.density..m2+ci), position = position_dodge()) +
  labs(title="Grazing impacts depend on nitrogen levels",
       x= "Grazing level",
       y= expression(paste("# of shoots/" ,m^{-2})))
```

### 4

Find an example of a factorial ANOVA from a paper that is related to
your work. Make sure you understand the connections between the methods,
results, and graphs. Briefly answer the following questions

-   What was the dependent variable?
-   What were the independent variables?
-   Was the interaction significant?
    -   If so, how did they interpret findings
    -   If not, were the main effects significant?
