---
title: Combining (lots of) numerical and categorical predictors
subtitle: ANCOVAs and beyond
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* use visual mode or **render** your file to produce a version that you can see!
* **render** your file to make sure it runs (and that you haven't been working
out of order)
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Example

Following the iris example from class

```{r}
set.seed(3)
iris_example_species <-data.frame(
  Species = c(rep("baruch",25), rep("hunter", 25), rep("york", 25)),
  Petal_Length = runif(75,2,4 ))
set.seed(31)
iris_example_species$Sepal_interaction <- 
  iris_example_species$Petal_Length * c(rep(-2, 25),rep(0,25), rep(5,25)) + 
  c(rep(2,25), rep(3,25), rep(4,25)) + rnorm(75)
```

Plot the data

```{r}
library(ggplot2)
ggplot(iris_example_species, aes(x= Petal_Length, y = Sepal_interaction, color = Species)) +
  geom_point()+
  ylab("Sepal Length") +
  xlab("Petal Length") +
  ggtitle("Impact of petal length and species on sepal length") +
    geom_smooth(method = "lm", se = F)
```

Analysis would indicate (assumption plots not shown here to allow focus
on interpreting interactions)

```{r}
library(car)
Anova(lm( Sepal_interaction~ Petal_Length * Species, iris_example_species), 
      type = "III")
```

interactions do exist. This means we can't interpret the "general"
relationship, so we need to look for each species using regression.

```{r}
summary(lm(Sepal_interaction ~ Petal_Length, 
         iris_example_species[iris_example_species$Species == "baruch",]))
Anova(lm(Sepal_interaction ~ Petal_Length, 
         iris_example_species[iris_example_species$Species == "baruch",]), 
      type="III")
summary(lm(Sepal_interaction ~ Petal_Length, 
           iris_example_species[iris_example_species$Species == "hunter",]))
Anova(lm(Sepal_interaction ~ Petal_Length, 
         iris_example_species[iris_example_species$Species == "hunter",]), 
      type="III")
summary(lm(Sepal_interaction ~ Petal_Length, 
           iris_example_species[iris_example_species$Species == "york",]))
Anova(lm(Sepal_interaction ~ Petal_Length, 
         iris_example_species[iris_example_species$Species == "york",]), 
      type="III")
```

Here we see that there is a significant negative relationship (F~1,23~ =
50.36, p<0.001) between sepal and petal length for *I. baruch*, a
significant positive relationship (F~1,23~ = 269.53, p<0.001) between
sepal and petal length for *I. york*,and no relationship (F~1,23~ =
1.63, p<-0.21) between sepal and petal length for *I. hunter*.


## Practice

### 1
  

Data on FEV (forced expiratory volume), a measure of lung function, can
be found at 

http://www.statsci.org/data/general/fev.txt

More information on the dataset is available at 

http://www.statsci.org/data/general/fev.html.

Does the impact of age on FEV differ among genders? Consider how your answer to 
this differs from the previous assignment!

```{r}
fev <- read.table("http://www.statsci.org/data/general/fev.txt", header = T, 
                  stringsAsFactors = T)
fev_age <- lm(FEV ~ Age*Sex, fev)
plot(fev_age)
library(car)
Anova(fev_age, type = "III")
summary(fev_age)
```

*We can explore this question using an ANCOVA since the response is continuous and
the explanatory variables combine a categorical and continuous variable. Analysis of
residuals indicates the assumptions are met (no pattern, normal distribution). 
There is a significant interaction among age and gender on FEV (F~1,650~=64.535, 
p<.001).  We should explore impacts of age on each gender separately.*

```{r}
fev_age <- lm(FEV ~ Age, fev[fev$Sex == "Male",])
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```
*Age has a significant (F~1,334~ = 641, p < 0.01) positive 
(.27 L yr^-1^) impact on FEV in males.*

```{r}
fev_age <- lm(FEV ~ Age, fev[fev$Sex == "Female",])
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```
*Age also has a significant (F~1,316~ = 380, p < 0.01) positive 
(.16 L yr^-1^) impact on FEV in females. The interaction is likely
due to the higher rate of increase of FEV with age in males.*

```{r}
library(ggplot2)
ggplot(fev, aes(x=Age, y=FEV, color = Sex, shape = Sex)) +
  geom_point(size = 3) +
  ylab("FEV (L)") +
  ggtitle("FEV increases faster \n with age in males")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32)) +
    geom_smooth(method = "lm", se = F)
```


### 2

Data on home gas consumption at various temperatures before and after new insulation was installed has been collected @ 

http://www.statsci.org/data/general/insulgas.txt

More information on the data is available @

http://www.statsci.org/data/general/insulgas.html

Is there any relationship between these factors?  How would you test this,
and what type of plot would you produce to accompany your analysis?

```{r}
heat <- read.table("http://www.statsci.org/data/general/insulgas.txt", 
                   header = T, stringsAsFactors = T)
head(heat)
heat_model <- lm(Gas ~ Insulate * Temp, heat)
plot(heat_model)
require(car)
Anova(heat_model, type = "III")

ggplot(heat, aes_string(x="Temp", y="Gas", color = "Insulate")) +
  geom_point(size = 3) +
  ylab(expression(paste("Gas (1000 ",ft^3, ")")))+
  xlab(expression(paste("Temperature (", degree~C, ")")))+
  geom_smooth(method = "lm", se = F) +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))


```
*There is a significant relationship between insulation type (before/after) and
temperature on gas usage (F~1,40~=10.39, p<.01). Graphical analysis indicates the old (before) insulation led to higher overall gas usage and gas usage increased faster with colder temperature
compared to the new insulation. Statistical analysis bears this out*

```{r}
heat_model_old <- lm(Gas ~ Temp, heat[heat$Insulate == "Before",])
plot(heat_model_old)
summary(heat_model_old)
Anova(heat_model_old, type = "III")

heat_model_new<- lm(Gas ~ Temp, heat[heat$Insulate == "After",])
plot(heat_model_new)
summary(heat_model_new)
Anova(heat_model_new, type = "III")
```
*There is a significant relationship between gas usage and temperature for old 
and new insulation homes. However, the old insulation led to using 400 ft^3^ 
more gas per week to heat the 
house with every degree
drop in temperature, while the new insulation leads to a increase of only 
250 ft^3^ more gas per week with each degree drop.*  

### 3

Data on the height, diameter, and volume of cherry trees was collected for
use in developing an optimal model to predict timber volume.  Data is available @ 

http://www.statsci.org/data/general/cherry.txt

Use the data to justify an optimal model.

```{r}
cherry <- read.table("http://www.statsci.org/data/general/cherry.txt",
                     header = T)
head(cherry)

#if only considering main effects (one option)
cherry_full <- lm(Volume ~ Diam + Height, cherry)
plot(cherry_full)
library(car)
Anova(cherry_full, type = "III")
#both are significant, so finished

#could also consider interactions 
cherry_full <- lm(Volume ~ Diam * Height, cherry)
plot(cherry_full)
Anova(cherry_full, type = "III")
summary(cherry_full)
#all significant, so finished
```
*I used multiple regression to consider the impacts of both continuous and 
categorical explanatory variables on timber volume.  I used a top-down approach 
focused on p-values (F tests) in this example.  Both diameter and height (and their
interaction) are significant, so the full model is justified by the data. It 
explains 97.5% of the variation in volume.  AIC methods lead to a similar outcome*

```{r}
library(MASS)
stepAIC(cherry_full)
```

### 4

Over the course of five years, a professor asked students in his stats class 
to carry out a simple experiment.  Students were asked to measure their pulse 
rate, run for one minute, then measure their pulse rate again.  The students 
also filled out a questionnaire.  Data  include:

Variable | Description
:-:  | :-:
Height | Height (cm)
Weight | Weight (kg)
Age    | Age (years)
Gender | Sex (1 = male, 2 = female)
Smokes | Regular smoker? (1 = yes, 2 = no)
Alcohol | Regular drinker? (1 = yes, 2 = no)
Exercise | Frequency of exercise (1 = high, 2 = moderate, 3 = low)
Change | Percent change in pulse (pulse after experiment/pulse before experiment)
Year | Year of class (93 - 98)

Using the available data (available at 

https://docs.google.com/spreadsheets/d/e/2PACX-1vToN77M80enimQglwpFroooLzDtcQMh4qKbOuhbu-eVmU9buczh7nVV1BdI4T_ma-PfWUnQYmq-60RZ/pub?gid=942311716&single=true&output=csv )

determine the optimal subset of explanatory variables that should be used to
predict change pulse rate (Change) (focusing on main effects only, no 
interactions) and explain your choice of methods.  Interpret your results. Make
sure you can explain any changes you needed to make to the dataset or steps you 
used in your analysis.

```{r}
pulse_class_copy <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vToN77M80enimQglwpFroooLzDtcQMh4qKbOuhbu-eVmU9buczh7nVV1BdI4T_ma-PfWUnQYmq-60RZ/pub?gid=942311716&single=true&output=csv", stringsAsFactors = T)
pulse_class_copy$Gender <- as.factor(pulse_class_copy$Gender)
pulse_class_copy$Smokes <- as.factor (pulse_class_copy$Smokes)
pulse_class_copy$Alcohol <- as.factor(pulse_class_copy$Alcohol)

pulse_full <- lm(Change ~ ., pulse_class_copy )
pulse_final <- step(pulse_full)
#consider assumptions
plot(pulse_final)
Anova(pulse_final, type = "III")
summary(pulse_final)


#or
library(MuMIn)
options(na.action = "na.fail")
auto <- dredge(pulse_full)
library(rmarkdown)
paged_table(auto)
options(na.action = "na.omit")

```
*I used step based approach (which requires nested models) and large search method
above. Using the step approach only height and alcohol usage are retained in the 
final model, which explains 10% of the variation in pulse change. Model assumptions 
are also met. The search method finds the same optimal model but notes many other
models (including a null model) perform similarly well.*



### 5

Find one example of model selection from a paper in your field. It may be more
complicated (see next question!) than what we have done, but try to identify the 
approach (F/AIC, top-down/bottom-up/not nested) they used. Review how they 
explained their approach (methods) and reported outcomes (results). Be prepared
to discuss in class next week.

### 6 

Find one example of a linear model selection (e.g., generalized linear models, 
mixed-effects models, beta regression) from a paper in your field. Be prepared
to name the technique in class next week.