---
title: Acquiring data
subtitle: Sampling and related issues
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

## How do we get data?

Let's start our statistics journey by thinking about the simplest
scenario: We want to know something about a group. An example might be
the average (also known as the mean, we will define later if needed!),
the minimum value, or the maximum value for some trait. These questions
all deal with the distribution of values for that trait in the group.
These specific traits of the group distribution are called *statistics:*

> the numerical facts or data themselves - *Dictionary.com*

This means we have a target trait we are focused on, and we have defined
a group of interest. We can call this group of interest a *population*.
Note that while the term *population* may have specific meanings in some
fields (such as ecology), here population is just the group of interest.
It could be a population of Goliath grouper in Florida, a population of
flowers in Virginia, or people from a certain country or demographic
group. We could want to know something about all of these groups!

As we've already noted, in a perfect world we know everything (or at
least everything about our trait value) for every member of the focal
population. However, we often don't and can't measure every member of a
population. It may be too difficult or expensive to measure every member
of the population. In fact, we may not even know how large the
population is!

In the cases where we can't measure every member of the population, we
collect data on the focal trait(s) from a *sample*. A sample is the
subset of the population of interest. Data can be collected from samples
used in experimental studies, where researchers manipulate something to
see how it impacts the focal trait. Researchers may expose organisms to
different stimuli in controlled lab, field, or mesocosm studies to see
what happens. For example, researchers interested in impacts of an
invasive crayfish (*Pacifastacus leniusculus*) on Mazama newts (*Taricha
granulosa mazamae*) collected newts and crayfish.; they then placed
either just newts or newts and crayfish in in large tanks to observe
interactions [@girdner2018].

[![Experimental mesocosms used to evaluate Mazama newt and signal
crayfish behavior on Wizard Island, Crater Lake, Oregon. A team of NPS
scientists observed the interaction between newts and crayfish in tanks
designed to mimic natural habitat. NPS photo, public
domain.](https://www.nps.gov/common/uploads/stories/images/nri/20160414/articles/E3D75BF5-1DD8-B71B-0BAA0D9A58F1307A/E3D75BF5-1DD8-B71B-0BAA0D9A58F1307A.jpg){#fig-nps_mesocosm
fig-alt="Image of people filling large black tanks on dock near lake."
fig-align="center"}](https://www.nps.gov/articles/parkscience32_1_5-12_buktenica_et_al_3815.htm)

Data can also be collected from observational studies, where researchers
"simply" measure outcomes and other traits without manipulating
anything. For example, scientists interested in impacts of climate
change on species ranges surveyed sites for species presence and
abundance and compared it to historical data [@sagarin1999]. Note doing
this may actually be very difficult! @metcalfe2022 found that just
observing capuchin monkeys (*Sapajus macrocephalus*) changed their
behavior.

[![Large-headed capuchins studied by Metcalfe et al (2022) changed their
behavior based on observer density and distance. Cody H., CC BY-SA 2.0
\<https://creativecommons.org/licenses/by-sa/2.0\>, via Wikimedia
Commons.](/images/Large-headed_capuchin_(Sapajus_macrocephalus)_1.jpg){fig-alt="A monkey peers into distance while sitting in a tree."
fig-align="center"}](https://commons.wikimedia.org/wiki/File:Large-headed_capuchin_(Sapajus_macrocephalus)_1.jpg)

Different types of studies change what we can use the data for. We'll
come back to this later, but, in general, experimental studies are more
commonly used to ascertain causation (something makes something happen),
whereas observational studies are used to assess correlation (something
happens when something else happens, also known as association).
However, these can be hard to disentangle, especially since studies can
only be observational since experiments would be unethical or impossible
to carry out. As XKCD puts it

[![XKCD: Correlation. Title text (text that pops up when you hover over
the comic): Correlation doesn't imply causation, but it does waggle its
eyebrows suggestively and gesture furtively while mouthing 'look over
there'. https://xkcd.com/552/, CC BY-NC 2.5
\<https://creativecommons.org/licenses/by-nc/2.5/\>.](https://imgs.xkcd.com/comics/correlation.png){#fig-XKCD552
fig-alt="Picture of two stick figures chatting.  One says, \"I used to think correlation implied causation. Then I took a statistics course. Now I don't.\" The other figure responds \"Sounds like the class helped.\", To which the first figure responds, \"Maybe.\""
fig-align="center"}](https://xkcd.com/552/)

> Correlation doesn't imply causation, but it does waggle its eyebrows
> suggestively and gesture furtively while mouthing 'look over there' -
> XKCD #552

Once we have the sample, we can measure the trait of interest in it, and
use that to estimate the statistic of interest for the actual
population. This is the science of statistics, which can actually be
defined as

> the practice or science of collecting and analyzing numerical data in
> large quantities, especially for the purpose of inferring proportions
> in a whole from those in a representative sample. - *Oxford English
> Dictionary*

If the whole idea of statistics is to infer something about the
population from our sample, we need to make sure the sample is
*representative* of the population. That means it should not be
*biased*. *Bias* occurs if the trait values we measure in our sample
differ from the population in a consistent way. This can happen with
samples of convenience, or when researchers select samples that are easy
to measure but may not be representative of the population. Classic
examples include estimating the amount of time students spend studying
by surveying students at a campus library.

Bias may also be related to issues of independence. In a good sampling
design, every member of the population has the same chance of being
included in a sample. Samples of convenience violate this premise, and
often the underlying issue is that the samples are not independent. A
perfect solution is to randomly choose members of the population to be
in the sample, but that is often not possible. Again, it requires
knowing every member of the population! Independence also means each
data point is not related to any others!

[![XKCD: Slope Hypothesis Testing. Don't worry, we'll come back to
significance - but what is the independence issue?
https://xkcd.com/2533/, , CC BY-NC 2.5
\<https://creativecommons.org/licenses/by-nc/2.5/\>.](https://imgs.xkcd.com/comics/slope_hypothesis_testing_2x.png){#fig-XKCD2533
fig-alt="Comic strip where 2 figures discuss relationship between how loud students scream and exam grades."
fig-align="center"}](https://xkcd.com/2533/)

In some cases linkages among samples are impossible to avoid. We will
cover ways to address that using blocking factors or random effects
later.

Notice in discussing bias this way we are not directly focusing on the
quality of the measurements. Obviously we need good data to make good
estimates, but these ideas are different from our current focus on
picking a good sample. If we want to discuss the quality of our
measurements, we could think about accuracy (how well we measure the
underlying trait in regards to its true value, which we typically don't
know) and precision (how repeatable our measurement technique is).

[![Example of accuracy and precision. From sketchplantations.com,
https://sketchplanations.com/accuracy-and-precision, CC BY-NC 4.0
\<https://creativecommons.org/licenses/by-nc/4.0/\>](/images/accuracy_precision.png){fig-alt="Image shows four targets and states accuracy and precision are not the same thing.  The first target shows a wide spread of dots on target (not near each other or center); this is not accurate or precise. The second shows dots close to center of target but all around; this is accurate but not precise. The third target has a tight cluster of dots near one edge; this is precise but not accurate. The final target shows a tight cluster of dots in the center of the targer; this is precise and accurate."}](https://sketchplanations.com/accuracy-and-precision)

Even if we have a proper way to measure a trait (accurate and precise)
in a good sample (not biased), we will still be producing an estimate of
the population statistic! This is due to *sampling error*. Sampling
error refers to the fact that every sample will produce a slightly
different estimate of the statistic. Imagine this - there a 1000 fish in
a lake. We sample 50 of them, measure their length, and use it calculate
the average fish length. If we took a different sample, do you think it
would have exactly the same average?

We can demonstrate this in R - you won't understand the code below yet,
so just trust me for now, but this will let you start seeing code and
thinking about how to use it.

Let's generate a population of fish. We'll store their lengths in a
vector called *lengths*.

```{r}
lengths <- rnorm(n=1000, mean = 10, sd=1)
```

The average length of fish in this population is
`r round(mean(lengths),2)` cm (Note: if you view this on the webpage you
will see a number, but in the actual qmd file you see R code here - this
is an example of merging code and text!). We can then simulate a sample
from this population. In fact, let's simulate 2 and compare the means of
each.

```{r}
sample_1 <- sample(lengths,50)
sample_2 <-sample(lengths, 50)
```

The mean length for fish in sample 1 is `r round(mean(sample_1),2)` cm,
while that in sample 2 is `r round(mean(sample_2))` cm . These are both
close to the true value, but they are also both slightly different -
this is sampling error!

Sampling error always exists, and a major part of statistics is to
quantify it. One thing that reduces sampling error is to have large
samples! Remember, if we measure every member of the population we don't
even need statistics, so the closer we get to that (implying larger
samples) the better!

## Next steps

Now that we have data, we'll discuss summarizing it in the next section
(and actually define *mean* and some of the other terms we've started to
use!).
