---
title: Combining (lots of) numerical and categorical predictors
subtitle: ANCOVAs and beyond
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

So far we've used linear models to consider how categorical predictors
and continuous numerical predictors can predict continuous outcomes.
We've already considered having multiple categorical predictors
(remember factorial ANOVAs). Now we'll extend those ideas to models
where we have numerous categorical and/or continuous numerical
predictors in the same model. In doing so we'll introduce (and connect)
ANCOVAs, multiple regression, and model selection.

## Back to the iris data

We will motivate this (again!) with an example from our favorite iris
data. So far we have considered how species impacts sepal lengths
(abbreviated analysis:

```{r}
library(Rmisc)
function_output <- summarySE(iris, measurevar="Sepal.Length", groupvars =
                               c("Species"))

library(ggplot2)
ggplot(function_output, aes(y=Sepal.Length, x=Species, fill=Species)) +
  geom_col(aes(fill=Species)) +
    geom_errorbar(aes(ymin=Sepal.Length-ci, ymax=Sepal.Length+ci)) +
  labs(title=expression(paste("Sepal lengths of ",italic("I. species"))),
       y= "Sepal length (cm)",
       x= "Species")
```

```{r}
iris_anova <- lm(Sepal.Length~Species, iris)
plot(iris_anova)
library(car)
Anova(iris_anova, type="III")
library(multcomp)
compare_cont_tukey <- glht(iris_anova, linfct = mcp(Species = "Tukey"))
summary(compare_cont_tukey)
```

We've also considered how petal length influences sepal length
(abbreviated analysis:

```{r}
ggplot(iris, aes(y=Sepal.Length, x=Petal.Length)) +
  geom_point() +
  labs(title="Sepal length increases with petal length",
       y= "Sepal length (cm)",
       x= "Petal Length (cm)") +
  geom_smooth(method = "lm",se = F)

```

```{r}
iris_regression <- lm(Sepal.Length ~ Petal.Length, iris)
plot(iris_regression)
Anova(iris_regression, type = "III")
summary(iris_regression)
```

However, what if we wanted to consider the combined (and potentially
interacting) effects of petal length and species on sepal length? This
is like our 2-way ANOVA but with one predictor being continuous. This is
often called an ANCOVA, but it's just another linear model! Overall, we
are decomposing the variance of each data point among various factors.
Our use of type III residuals let's us ask how much any given factor
explains *given* that other factors are in the model (we'll explore this
more below). Put another way, we might want to know if a factor adds
explanatory power to the model (and is not redundant or subsumed by
another factor).

Continuing to use focal iris dataset, we can use the same format we used
for factorial or 2-way ANOVAs to add the factors.

```{r}
iris_ancova <- lm(Sepal.Length~Species*Petal.Length, iris)
```

This model includes an interaction and matche the following null
hypotheses (note 2 are copied from previous sections!):

$$
\begin{split}
H_O: \mu_{sepal \ length, \ setosa} = \mu_{sepal \ length, \ virginica} = \mu_{sepal \ length, \ versicolor}\\ 
H_O: \beta_\textrm{(coefficient between sepal and petal length)} = 0\\\\
H_O: \textrm{relationship between petal length and sepal length does not differ
among species}\\
\end{split}
$$ Once we develop the model, we can visually check the assumptions,
which remain

$$
\epsilon \approx i.i.d.\ N(\mu,\sigma)
$$

```{r}
plot(iris_ancova)
```

If assumptions appear to be met (as they do here), we can consider
impacts of factors. Given the presence of categorical predictors, an
Anova table may be informative

```{r}
Anova(iris_ancova, type="III")
```

Since the interaction is not significant (F~2,144~ = 1.68, p = 0.19), we
have the same 2 options we noted in factorial ANOVAs.

### Interpret results with the interaction

We can read results from the full model. These show no interaction
between species and petal length (F~2,144~ = 1.68, p = 0.1895) and no
impact of petal length (F~1,144~=3.83, p = 0.052) but a significant
impact of species (F~2,144~ = 12.81, p \< 0.01). We can follow this up
with a post-hoc test

```{{r}}
compare_ancova_tukey <- glht(iris_ancova, linfct = mcp(Species = "Tukey"))
summary(compare_ancova_tukey)
```

which suggests *I. versicolor* and *I. virginica* are similar to each
other but different from other species.

However, note the *glht* output notes interactions are still present in
model, so this approach may be inappropriate. For ANOVA's we noted could
instead compare means for each combination.

```{r}
library(emmeans)
emmeans(iris_ancova, pairwise ~ Species*Petal.Length)
```

This approach, however, is less useful for numerical predictors. By
default the **emmeans** package uses the average for the covariate
(note,

```{r}
mean(iris$Petal.Length)
```

is why we see *3.578* spread throughout the output). While we can
specify other breaks, this highlights the difference in combining
cateorical variable impacts and combining effects of categorical and
continuous variables.

### Remove the interaction

Since the interaction isn't significant, we can also remove it.

```{r}
iris_ancova_updated <- lm(Sepal.Length~Species+Petal.Length, iris)
plot(iris_ancova_updated)
Anova(iris_ancova_updated, type="III")
```

This shows us that both petal length and species impact sepal length.

Why did these factors "suddenly" become significant. The issue lies in
that they are not fully independent. Note

```{r}
petal_anova <- lm(Petal.Length ~ Species, iris)
plot(petal_anova)
Anova(petal_anova, type ="III")
summary(petal_anova)
```

This shows that species explains a lot
(`r summary(petal_anova)$r.squared` %, in fact) of the variation in
petal length. Thus interactions among species and petal length are hard
to define, as each species petal lengths are *mostly* different than the
others.

This relates to an issue we will soon see. When you consider the impacts
of multiple factors on a variable, you are assuming they are not
related. That is rarely true (except for when we set up factorial
ANOVAs), so we have to decide how related is ok. As factors become more
related, the linear model approach does not work.On one level, its hard
to split variance correctly among two similar columns. Mathematically,
it also makes the design matrix harder to invert.

<details>

<summary>How does this relate to types of residuals?</summary>

This also relates to the "types" of residuals. Type 1 notes the order of
factors in the model. The function *anova* uses this type. Note the
values are different (but still significant) depending on the order.

```{r}
anova(iris_ancova)
```

and

```{r}
iris_ancova_reversed <- lm(Sepal.Length~Petal.Length*Species, iris)
anova(iris_ancova_reversed)
```

This is because Type 1 residuals remove all the variation that can be
explained by the first factor (even if it *could* be given to a related
factor), then do the same for the second and so forth. So related
factors will have different p-values depending on order (unbalanced
designs also impact this.)

Type II residuals don't focus on order, but they ignore interactions. In
doing so they ignore any variation that could be attributed to multiple
factors.

Type III residuals include interactions, but ask how much a given factor
contributes to the explanatory power of a model given main effects of
other factors and interactions, including those with the focal factor,
are already present. This use of *marginal means* may seem odd (e.g.,
asking if a factor should be included when you are already including
interactions with said factor) and also means the sum of squares that we
get from decomposing the variance adds up to *more* than the total sum
of squares. However, from a conceptual standpoint they work and are thus
commonly used.

</details>

### What would interactions look like?

So, what would interactions actually look like, and how would you
interpret them? To illustrate this, let's pretend we visit another
valley and sample 3 new iris species (*I. baruch*, *I. hunter*, and *I.
york)*. We want to see how species and petal length impact sepal length
in these species. Let's make some data to suggest potential outcomes.

```{r}
set.seed(3)
iris_example_species <-data.frame(
  Species = c(rep("baruch",25), rep("hunter", 25), rep("york", 25)),
  Petal_Length = runif(75,2,4 ),
  #no difference based on species or sepal length
  Sepal_no_impacts = runif (75, 4, 6))
iris_example_species$Species <- factor(iris_example_species$Species)
```

First, we might find no relationship between the variables.

```{r}
ggplot(iris_example_species, aes(x= Petal_Length, y = Sepal_no_impacts, color = Species)) +
  geom_point()+
  ylab("Sepal Length") +
  xlab("Petal Length") +
  ggtitle("Impact of Petal length and species on sepal length") +
    geom_smooth(method = "lm", se = F)

```

Analysis would indicate (assumption plots not shown here to allow focus
on interpreting interactions)

```{r}
Anova(lm( Sepal_no_impacts~ Petal_Length * Species, iris_example_species), 
      type = "III")
```

no impact of interaction, so we could drop it

```{r}
Anova(lm( Sepal_no_impacts~ Petal_Length + Species, iris_example_species), 
      type = "III")
```

but we still find no main effects.

Alternatively, we might find a situation like this

```{r}
set.seed(77)
iris_example_species$Sepal_no_impact_species <- 
  iris_example_species$Petal_Length * 2 + rnorm(75)

ggplot(iris_example_species, aes(x= Petal_Length, y = Sepal_no_impact_species, color = Species)) +
  geom_point()+
  ylab("Sepal Length") +
  xlab("Petal Length") +
  ggtitle("Impact of Petal length and species on sepal length") +
    geom_smooth(method = "lm", se = F)

```

Analysis would indicate (assumption plots not shown here to allow focus
on interpreting interactions)

```{r}
Anova(lm( Sepal_no_impact_species~ Petal_Length * Species, iris_example_species), 
      type = "III")
```

no impact of interaction, so we could drop it

```{r}
Anova(lm( Sepal_no_impact_species~ Petal_Length + Species, iris_example_species), 
      type = "III")
```

and note only petal length impacts sepal length.  

We could also ssee a situation like this

```{r}
set.seed(79)
iris_example_species$Sepal_no_relationship_petal <-rnorm(75) +
  c(rep(2,25), rep(4,25), rep(6,25))

ggplot(iris_example_species, aes(x= Petal_Length, y = Sepal_no_relationship_petal, color = Species)) +
  geom_point()+
  ylab("Sepal Length") +
  xlab("Petal Length") +
  ggtitle("Impact of Petal length and species on sepal length") +
    geom_smooth(method = "lm", se = F)

```

Analysis would indicate (assumption plots not shown here to allow focus
on interpreting interactions)

```{r}
Anova(lm( Sepal_no_relationship_petal~ Petal_Length * Species, iris_example_species), 
      type = "III")
```

no impact of interaction, so we could drop it

```{r}
Anova(lm( Sepal_no_relationship_petal~ Petal_Length + Species, iris_example_species), 
      type = "III")
```

and we find only species impacts sepal length. In that case, we need a
post-hoc follow up.

```{r}
summary(glht(lm( Sepal_no_relationship_petal~ Petal_Length + Species, iris_example_species), 
             linfct =mcp(Species = "Tukey")))
```




## Next steps
