---
title: More ANOVAs
subtitle: Dealing with multiple group membership and interactions
bibliography: ../references.bib
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")
```

In the last chapter we introduced the idea of comparing means among
populations (one-way ANOVAs, our first linear models). However, the
units that we measure may belong to multiple groups. We will extend our
analysis of variance to consider multiple group membership and
interactions in this chapter. As a starting point, consider that group
membership may be an inherent property of the unit we measure or we may
assign it.

## Example: Back to the birds

One of the last chapters practice problems focused bird feathers. While
studying feather color in Northern flickers (*Colaptes auratus*),
@wiebe2002 noted that \~25% of birds had one or more "odd" tail
feathers. They decided to compare the color of these odd and "typical"
feathers.

[![Northern Flicker. Mike's Birds, CC BY-SA 2.0
\<https://creativecommons.org/licenses/by-sa/2.0\>, via Wikimedia
Commons](/images/Northern_Flicker_(4508135578).jpg)](https://commons.wikimedia.org/wiki/File:Northern_Flicker_(4508135578).jpg)

Example and data provided by @mcdonald2014.

```{r, echo=FALSE}
Input = ("
 Bird    Feather_type   Color_index
 A       Typical   -0.255
 B       Typical   -0.213
 C       Typical   -0.19
 D       Typical   -0.185
 E       Typical   -0.045
 F       Typical   -0.025
 G       Typical   -0.015
 H       Typical    0.003
 I       Typical    0.015
 J       Typical    0.02
 K       Typical    0.023
 L       Typical    0.04
 M       Typical    0.04
 N       Typical    0.05
 O       Typical    0.055
 P       Typical    0.058
 A       Odd       -0.324
 B       Odd       -0.185
 C       Odd       -0.299
 D       Odd       -0.144
 E       Odd       -0.027
 F       Odd       -0.039
 G       Odd       -0.264
 H       Odd       -0.077
 I       Odd       -0.017
 J       Odd       -0.169
 K       Odd       -0.096
 L       Odd       -0.33
 M       Odd       -0.346
 N       Odd       -0.191
 O       Odd       -0.128
 P       Odd       -0.182
")

feather <-  read.table(textConnection(Input),header=TRUE)
```

```{r}
library(rmarkdown)
paged_table(feather)
```

## How do we analyze this data?

We may first note that we have a continuous measurement (feather color,
measured using color hues from a digital camera and another statistical
technique that we will not go into here) and a categorical variable
(feather type, with levels "typical" and "odd"). This hopefully reminds
you of an ANOVA/t-test!

We could plot the data

```{r}
library(ggplot2)
ggplot(feather, aes(x=Feather_type, y= Color_index, color=Feather_type))+
  geom_jitter()+
  labs(y= "Color index",
       x= "Feather type",
       title="Comparing odd and typical feathers in Northern flickers")+
  guides(color=F)
```

Develop a set of hypotheses:

$$
\begin{split}
H_O: \mu_{\textrm{odd feather color}} = \mu_{\textrm{typical feather color}}\\
H_A: \mu_{\textrm{odd feather color}} \neq \mu_{\textrm{typical feather color}}\\
\end{split}
$$

and test them using a t-test:

```{r}
t.test(Color_index ~ Feather_type, data=feather)
```

or, using more generalizable functions, a linear model:

```{r}
library(car)
Anova(lm(Color_index ~ Feather_type, data=feather), type = "III")
```

We find a significant p value, **but we did not check assumptions**. For
linear models (remember, \$\epsilon \approx i.i.d. N(\mu,\sigma)\$, we
could use our visual checks

```{r}
plot(lm(Color_index ~ Feather_type, data=feather))
```

Which appears ok, but there is a problem.

Our data are not independent!

## Lack of Independence

Odd and typical feathers were measured on a single bird (note the *Bird*
column) in the dataset. We might assume feathers on a given bird are
more closely related in color than feathers on different birds. This
could be due to diet or other factors making all feathers on a given
bird brighter or darker than those on another. Regardless of reason (and
"good" p value), we know the measurements are linked in some way. Note
we could "connect" individual observations.

```{r}
ggplot(feather, aes(x=Feather_type, y= Color_index, color=Feather_type, group=Bird))+
  geom_line(position = position_dodge(0.4), color="black") +
  geom_point(position = position_dodge(0.4)) +  
  labs(y= "Color index",
       x= "Feather type",
       title="Comparing odd and typical feathers in Northern flickers")+
  guides(color=F)
```

This may also occur if we measure outcomes with-in a single unit (e.g.,
a study of fertilizer impacts using multiple fields) or over time (e.g.,
before/after studies). Regardless of the reason, when our experimental
design has led to measurements being connected/not independent, we need
to consider these connections in order to properly note (and sometimes
even observe) impacts of focal variables.

## Blocking, two-way ANOVAs,and paired t-tests

In this case, the connections may be considered artifacts of the data.
We didn't assign birds. We also made a choice to compare odd and typical
feathers from the same bird - why? In general, accounting for extra
variation in the data will give you a better answer about how a given
variable influences outcomes. This may be called *blocking*. Although
the motivation might therefore be to get a "better" p value, it should
be driven by experimental design (and thus we started with an example
where we didn't "need" to account for it to achieve significance).

In order to consider how color differs by bird *and* feather type, we
need to add both variables to our linear model. For each variable we
add, we also add a null (and corresponding alternative) hypothesis. So
we retain our focus on feather type:

$$
\begin{split}
H_O: \mu_{\textrm{odd feather color}} = \mu_{\textrm{typical feather color}}\\
H_A: \mu_{\textrm{odd feather color}} \neq \mu_{\textrm{typical feather color}}\\
\end{split}
$$

but also add a set of hypotheses focused on birds:

$$
\begin{split}
H_O: \mu_{\textrm{color of bird A}} = \mu_{\textrm{color of bird B}}....\textrm{for all birds}\\
H_A: \mu_{\textrm{color of bird A}} \neq \mu_{\textrm{color of bird B}}....\textrm{for all birds}\\
\end{split}
$$

We can analyze this using our linear model approach. This is possible
because, as we noted earlier, we can subdivide variance among multiple
levels. Under the hood, the linear model approach build a model matrix
that considers the impact of feather type and bird on outcomes. Since
both variables are categorical, this is often called a two-way ANOVA.
First, let's make the object

```{r}
two_way_anova_example <- lm(Color_index ~ Feather_type + Bird, data=feather)

```

<details>

<summary>You can see the new model matrix and coefficients if you
want</summary>

Note the model matrix now includes columns for feather type and bird
(lots of dummy variables, and now intercept is Bird A's odd feather!).
The $\beta$ matrix of coefficients has corresponding estimates.

```{r}
library(rmarkdown)
paged_table(data.frame(model.matrix(two_way_anova_example)))
matrix(as.numeric(two_way_anova_example$coefficients), ncol=1)

```

So for our first observation, which is

```{r}
feather[1,]
```

Our estimate is the intercept (since it's bird A) and the typical
feather:

```{r}
model.matrix(two_way_anova_example)[1,] %*% matrix(as.numeric(two_way_anova_example$coefficients), ncol=1)

```

and thus our residual is

```{r}
two_way_anova_example$residuals[1]

```

which is the same as

```{r}
feather[1,]$Color_index-model.matrix(two_way_anova_example)[1,] %*% matrix(as.numeric(two_way_anova_example$coefficients), ncol=1)

```

</details>

Then check the assumptions

```{r}
plot(two_way_anova_example)
```

Note, visually speaking, the residuals do appear to be closer to normal
now. Since assumptions look ok, we can analyze the outcome

```{r}
summary(two_way_anova_example)
Anova(two_way_anova_example, type= "III")
```

Note we see a significant difference in color among birds and feather
type. Although we may be tempted to (and could) use post-hoc tests to
consider which birds are different than which others, this is typically
not done for blocked variables. We did not assign these pairings and it
is not the focus of our efforts.

Since we only had 2 types of feathers, we also don't need post-hoc
tests. A significant p value means they differ from each other, and the
estimates provided by the *summary* command indicate the typical
feathers have a higher color index.

### t-test connections

When we have only two measurements per group (e.g., odd and typical
feathers from each bird), we can use a t-test approach to achieve
similar goals. This approach is known as a paired t-test. Instead of
focusing on the difference in means (like a 2-sample t-test), the test
focuses on the mean difference between paired measurements (which would
be 0 under the null hypothesis!). In this way, it is effectively a
one-sample test that is pairing the data to reduce variation (blocking).
We can carry out the test, but (as of early 2024) we need to use a wide
dataset.

<!-- t.test no longer allowing paired in formula method as of early 2024  -->

```{r}
library(reshape2) 
feather_wide <- dcast(feather, Bird~Feather_type, value.var="Color_index")
paged_table(feather_wide)
```

Now we have wide data...

```{r}
t.test(feather_wide$Odd, feather_wide$Typical, data=feather, paired=TRUE)
```

and get the same results as above (note we don't even have to consider
corrections like the Welch approach since this a one-sample test).
Common examples of paired t-tests include before-after and twin studies.

In an earlier chapters we considered options for one- and two-sample
tests when t-tests assumptions were not met. For two-sample tests, one
of these approaches, the sign or binary test, is only valid for paired
data. The differences in paired observations are compared to a set value
(typically 0). Under the null hypothesis, half should be below the
proposed median and half should be above. Differences matching the
proposed value are ignored, thus reducing the sample size and making it
harder to reject the null hypothesis; this is actually an odd way of
accounting for them. The proportion of values below the proposed median
is then evaluated using a binomial test. For two sample, the *SIGN.test*
function in the **BSDA** package requires 2 columns of data and assumes
the order of the column represents paired data.

```{r}
library(BSDA)
SIGN.test(feather[feather$Feather_type == "Odd", "Color_index"], 
          feather[feather$Feather_type == "Typical", "Color_index"],
          md = 0)

```

### More than 2 measurements? Back to the linear model

We can also block for variation when we take more than 2 measurements
per unit. For example, imagine if these birds also had a special, long
tail feather.

```{r}
set.seed(25)
special <- data.frame(Bird = LETTERS[1:16], Feather_type = "Special", 
                      Color_index= feather[feather$Feather_type == "Typical", "Color_index"] +
                        .3 +runif(16,1,1)*.01)
feather_extra <- merge(feather, special, all = T)
feather_extra$Feather_type <- factor(feather_extra$Feather_type)
```

We could still block for variation using the linear model/ANOVA, but not
the t-test, approach. As another review, we create the model

```{r}
more_blocks <-lm(Color_index ~ Feather_type + Bird, data=feather_extra)
```

Check assumptions

```{r}
plot(more_blocks)
```

Check outcome (this time focusing on *Anova* output)

```{r}
Anova(more_blocks, type="III")
```

We still see feather type has a significant impact on color, but since
we have more than 2 groups we need to follow up this finding with a
post-hoc test.

```{r}
library(multcomp)
compare <- glht(more_blocks, linfct = mcp(Feather_type = "Tukey"))
summary(compare)
```

## Other ways to be in multiple groups

In the bird example, one of our categories (bird) was un-intential. We
chose to measure odd and typical feathers, and accounting for variation
among birds was an appropriate step given lack of independence in
measurements. However, we can also assign units to multiple groups
(instead of making multiple measurements within one unit).

Consider if we ran an experiment focused on the impact of factors A and
B on some trait. We can fully cross the factors in an experiment. Doing
so can let us consider the **main effects** of multiple variables and
potential **interactions** among them in what is often called a
**factorial ANOVA.** For starters, let each factor have only 2 levels,
and let the levels be the absence or presence of the factor.

|              |                  |                        |
|--------------|------------------|------------------------|
|              | **Factor A**     |                        |
| **Factor B** | *Absent*         | *Present*              |
| *Absent*     | Control          | Impact of A only       |
| *Present*    | Impact of B only | Combined impact of A+B |

<details>

<summary>Experimental design notes</summary>

For a factorial ANOVA, we need to assign each unit randomly to a level
of factor A. Then each level of factor B is randomly assigned to
subjects at each level of factor A. This is different than randomly
assigning treatments of A and B, as that could lead to outcomes where
some level of factor B is not represented in some level of factor A.

We also need multiple units (3+) assigned to each combination.

</details>

When both are absent we have a classic control outcome. When one is
present and the other absent we see **main effects** impacts of only one
factor. Note we previously analyzed experiments that considered *only*
one factor using ANOVAs or t-tests (linear models), but now we have
multiple factors. **We should not analyze the main effects of each using
2 one-way ANOVAs**. Doing so cuts our data in half, meaning our
estimates of variances are less precise and we increase our chance of
making a type 1 error. More importantly, we wouldn't be able to properly
consider the combined impacts of A + B. What could these be?

```{r}
example_interaction <- data.frame(Treatment = c(rep("Control",5),
                                                rep("Impact of A only",5),
                                                rep("Impact of B only",5),
                                                rep("A+B Additive",5), 
                                                rep("A+B Synergistic", 5),
                                                rep("A+B Antagonistic", 5)), 
                                  Cause= rep(c("Control","Factor A","Factor B", "Synergistic", "Antagonistic"), 6),
                                  Impact = c(5,0,0,0,0,
                                             5,2,0,0,0,
                                             5,0,3,0,0,
                                             5,2,3,0,0,
                                             0,0,0,20,0,
                                             0,0,0,0,6))
example_interaction$Treatment <- factor(example_interaction$Treatment, levels=c("Control","Impact of A only","Impact of B only", "A+B Additive", "A+B Synergistic", "A+B Antagonistic"))
example_interaction$Cause <- factor(example_interaction$Cause, levels=c("Control","Factor A","Factor B", "Synergistic", "Antagonistic"))
ggplot(example_interaction, aes(x=Treatment, y= Impact, fill= Cause))+
  geom_col(position = position_stack(reverse = TRUE))+
  theme(axis.text.x = element_text(angle = -45))
  
```

As shown in the graph (Inspired by [@fong2017]), A and B could have
additive effects (where they simply stack), synergistic effects (the
combined impact is more than the sum of the two), or antagonistic
effects (the combined impacts is less than the sum of the two).
Synergistic and antagonistic impacts are both examples of
**interactions**. Interactions occur when the impact of one variable
depends on the level of another.

### Example: Impacts of grazing and fertilization

We can extend this example to consider more than 2 levels for one or
more factors. For example, @valdez2023 wanted to consider the impact of
top-down (snail grazing) and bottom- up (nutrient availability) on marsh
plant (*Spartina alterniflora*) growth. To do this, they assigned plots
to one of 3 grazer treatments and one of 2 nitrogen treatments.

[![Fig 1 from Valdez et al. 2003. Map and conceptual illustration of
experimental
design.](/images/journal.pone.0286327.g001.PNG){fig-alt="Map of study site on Hog Island, Virginia, USA and conceptual illustration of experimental design with the following treatments: 1) Nitrogen addition with ambient snails, 2) nitrogen addition with three times ambient snails, 3) nitrogen addition without snails, 4) ambient nitrogen with ambient snails, 5) ambient nitrogen with three times ambient snails, and 6) ambient nitrogen without snails. The figure also depicts cage controls and uncaged plots used to assess caging effects on marsh plants. The map in the figure was created in R using ggmap [33] from ©OpenStreetMap under a ODb license, with permission from OpenStreetMapFoundation, original copyright 2018. https://www.openstreetmap.org/copyright."}](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0286327)

This design is different from the bird example. No two measurements for
a given trait were taken on the same plot. In this case, we likely care
about the main effects, or impacts, of both variables. However, we may
also need to consider interactions among the variables. Interactions
occur when the impact of one variable depends on the level of another.
For example, snail removal might have major impacts on nitrogen-enriched
plots while having no impact on ambient plots. Due to this, we now have
even more hypotheses:

$$
\begin{split}
H_O: \mu_\textrm{plant growth, no fertilizer} = \mu_\textrm{plant growth, fertilizer}\\
H_O: \mu_\textrm{plant growth, snails removed} = \mu_\textrm{plant growth, control snails}= \mu_\textrm{plant growth, snails added}\\
H_O: \textrm{impact of snail grazing does not depend on nitrogen level}\\
\end{split}
$$

Fortunately, these are easy to consider in our linear model framework.
While not shown here, the model matrix adds columns to note our new
interaction terms, and the coefficient matrix estimates them. From an R
standpoint, we can include the interaction between two variables using
the ":" notation. We'll focus on below-ground biomass (standardized to
m^2^) for this example (the paper measured 9 response variables!)

```{r}
valdez_2023 <- read.csv("data/Spartina_alterniflora_traits.csv", stringsAsFactors = T)
bgb_model <-lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023)
```

For shorthand, note that if we put *main effect \* main effect* in a
model, it automatically adds the interaction term. You can see the model
summary is the same.

```{r}
bgb_model_shorthand <-lm(below.biomass.g.meter.sq..m2..~Snail.Level * Nitrogen.level, valdez_2023)
summary(bgb_model)
summary(bgb_model_shorthand)
```

You may note a weird NA here (we'll come back to it), but remember we
should really check model assumptions before looking at output.

```{r}
plot(bgb_model)
```

These look ok. There may be a slight increase in variance with fitted
values, but we can work with this. Let's build an ANOVA table.

```{r, error = T}
Anova(bgb_model, type="III")
```

But we got an error! What happened? Let's look at the data

```{r}
paged_table(valdez_2023)
```

A summary may help more. Note we can summarize across multiple factors.

```{r}
library(Rmisc)
paged_table(summarySE(valdez_2023, measurevar = "below.biomass.g.meter.sq..m2..", groupvars = c("Snail.Level", "Nitrogen.level")))
```

Note the *uncaged* treatment only has *without* for the nitrogen impact.
It was a control! While we often need these in experiments, they can
create analysis problems. This is because we can't consider how nutrient
level depends on snail treatment for the control level! In other words,
interactions can not be calculated for some levels.

This is also why we saw the NA and warnings in our model summary

```{r}
summary(bgb_model)
```

You could note we have the same issue for our initial bird analysis:

```{r,error=T}
two_way_anova_example_int <- lm(Color_index ~ Feather_type * Bird, data=feather)
Anova(two_way_anova_example_int, type="III")
```

On a positive note, this means R will typically not consider
interactions when it shouldn't, but you need to know why in order to fix
it.

### Dealing with controls and missing interactions

To fix this (and deal with controls), we need to consider the data.
@valdez2023 used t-tests (why?) to consider differences between cage and
cage control plots (note %in% and the fact they did not focus on
above-ground biomass (maybe because uncaged plots had little..). %in%
allows you to subset data by matching items to list. Remember you can
always get help on functions using something like (we need the
quotations for operators!)

```{r, eval=F}
?'%in%'
```

```{r}
t.test(below.biomass.g.meter.sq..m2..~Snail.Level, valdez_2023[valdez_2023$Snail.Level %in% c("uncaged","control snails") & valdez_2023$Nitrogen.level == "without",])
```

If interactions are missing for other reasons (e.g., a set of units
failed/died/data was lost), we can either ignore interactions or combine
factor levels into a single new treatment variable and analyze using
one-way ANOVAs.

### Considering interactions

To consider interactions, we can remove the controls

```{r}
bgb_model_cont_removed <-lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",])
```

We can consider the assumptions

```{r}
plot(bgb_model_cont_removed)
```

and now note ...

```{r}
Anova(bgb_model_cont_removed, type="III")
```

...that the ANOVA table works.

Before we can interpret this, we need to return to our discussion about
different sums of squares.

### Back to type = III

Much of this follows an excellent post by . A paper by @venables may
also be helpful.

We introduced the various types of sums of squares in the [previous
chapter](Compare_means_among_populations.qmd), but they didn't matter
much there. However, here they do. This returns to the idea that we can
partition the variance of a given point among various factors.In
general, we can think about determining the impact by comparing sums of
squares from models with and without a factor of interest (using
*incremental* sums of squares). So, for a model

```         
lm(response ~ A + B + A:B)
```

We could consider the impact of all factors and interactions by
comparing:

```         
SS(AB | A, B) = SS(A, B, A:B) – SS(A, B)
SS(A | B, AB) = SS(A, B, A:B) – SS(B, A:B)
SS(B | A, AB) = SS(A, B, A:B) – SS(A, A:B)
SS(A | B) = SS(A, B) – SS(B)
SS(B | A) = SS(A, B) – SS(A)
```

However, the variance we attribute to each factor can be calculated in
multiple ways. First, we can consider type 1 sums of squares. Using this
approach, the order factors are entered into the model matters.
Following above,

```         
SS(A) for factor A. 
SS(B | A) for factor B. 
SS(A:B | B, A) for interaction A:B
```

This approach leads to different main effects (overall, general impacts
of a factor) for factors A and B depending on the order. Note

```{r}
summary(aov(bgb_model_cont_removed))
```

is different from

```{r}
summary(aov(lm(below.biomass.g.meter.sq..m2..~Nitrogen.level + Snail.Level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",])))
```

..Wait, it's the same. That's because differences only arise when the
data are unbalanced! ANOVAs were designed for use in perfectly balanced
datasets. We'll come back to this.

Another way is to consider the impact of main effects, such as

```         
SS(A | B) for factor A. 
SS(B | A) for factor B.
```

This approach is known as type II. It does not consider interactions (or
assumes they are equal to 0)

```{r}
Anova(bgb_model_cont_removed, type="II")
```

so it works ok here since the interaction is not significant, but it
shouldn't be used when interactions are significant/important. Compare
it to

```{r}
Anova(bgb_model_cont_removed, type="III")
```

to see the difference. Type III considers interactions.

To make things, more complicated, when looking at type III output, we
need to consider how R computes main effects. The default *contrast* in
R looks at the main effect for factor B using only the first level of
factor A (specified by the treatment contrasts). If there is an
interaction, this obviously doesn't make sense - B may matter for some
levels of A but not others. Given this, we can (and maybe should)
instead look at the *average* impact of B. We can do this by updating
how contrasts are calculated. We can specify overall changes to how R
does this, or we can include in the model as shown here:

```{r}
Anova(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum")), type = "III")
```

This time we get the same results for the main effects as we did for
Type 1 and Type 2 approaches. To be clear, this is because we used the
same type of contrasts and had a balanced design. Note this matches the
above,

```{r}
Anova(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum")), type = "II")
```

and this does not.

```{r}
Anova(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.treatment", Nitrogen.level = "contr.treatment")), type = "III")
```

In doing this, however, note our coefficients also change. Note

```{r}
data.frame(coef(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum"))))
```

is different than

```{r}
data.frame(coef(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",])))
```

The contrasts sum approach sets the intercept as the overall average and
then puts all other levels relative to that (meaning it includes all
levels). Our default treatment approach lets the first factor level
(typically alphabetical, here technically a grouped level (Snail Level
(ambient): Nitrogen level (with) ) be an intercept for the linear model,
and then considers the other factor levels as deviations from that.

Finally, note than unbalanced designs (when you have a different number
of means for one grouping than others), the different approaches lead to
different outcomes. We can consider this (and reinforce the above
points) using a built-in dataset, ToothGrowth. If you run this code

```{r, eval=F}
?ToothGrowth
```

in R, you'll see it notes the dataset contains "the length of
odontoblasts (cells responsible for tooth growth) in 60 guinea pigs.
Each animal received one of three dose levels of vitamin C (0.5, 1, and
2 mg/day) by one of two delivery methods, orange juice or ascorbic acid
(a form of vitamin C and coded as VC)."

You can view the data using:

```{r}
head(ToothGrowth)
```

and we can see we have a balanced design here:

```{r}
xtabs(~supp+dose, ToothGrowth)
```

Given that, our type 1, 2, 3 approaches are equivalent (we need to make
does a factor for ANOVA purposes)

```{r}
ToothGrowth$dose <- factor(ToothGrowth$dose) 
summary(aov(lm(len~supp*dose, ToothGrowth)))
```

```{r}
Anova(lm(len~supp*dose, ToothGrowth), type = "II") 
```

```{r}
Anova(lm(len~supp*dose, ToothGrowth, contrasts = list(supp = "contr.sum", dose = "contr.sum")), type = "III")
```

but not if we use a different contrast (treatment is default used here)

```{r}
Anova(lm(len~supp*dose, ToothGrowth), type = "III")
```

Also, if we have an unbalanced design

```{r}
library(purrr) 
library(tidyr) 
set.seed(19) 
ToothGrowth_unbalanced <- ToothGrowth %>%   group_by(supp, dose) %>%    nest() %>%               ungroup() %>%    mutate(n = c(10,8,7,5,10,3)) %>%    mutate(samp = map2(data, n, sample_n)) %>%    dplyr::select(-data) %>%   unnest(samp) 
ToothGrowth_unbalanced_anova <- lm(len~supp*dose, ToothGrowth_unbalanced, contrasts = list(supp = "contr.sum", dose = "contr.sum"))
```

Then we get different answers from each approach:

```{r}
summary(aov(ToothGrowth_unbalanced_anova))
```

```{r}
Anova(ToothGrowth_unbalanced_anova, type= "II")
```

```{r}
Anova(ToothGrowth_unbalanced_anova, type= "III")
```

Now we see the different ways of calculating sums of squares leads to
different outcomes when the sample sizes are unequal. In general, this
supports the use of the type III approach (but see @langsrud2003) and
for considering contrasts as needed.

### Interpreting interactions

An important note is that all approaches above found the same values (p,
F, etc.) for the highest order interaction. In general, interactions
make interpreting main effects difficult. So, we have a few ways of
handling them.

#### When not significant

If interactions are not significant, they can be handled in 2 ways.

1.  We can leave the interaction in the model and interpret main effects
    immediately

    1.  If doing this, we should make sure we set the contrasts to use
        an average approach (usually using the *sum* contrasts as shown
        above)

    ```{r}
        Anova(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level + Snail.Level:Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum")), type = "III")
    ```

2.  We can remove the interaction from the model, re-run it, and
    interpret main effects of the factors.

    ```{r}
    bgb_model_cont_removed_int_removed <- update(bgb_model_cont_removed, .~.-Snail.Level:Nitrogen.level)
    plot(bgb_model_cont_removed_int_removed)
    Anova(bgb_model_cont_removed_int_removed, type="III")
    ```

Which is the same as...

```{r}
Anova(bgb_model_cont_removed_int_removed, type="II")
```

When we do this, the contrasts methods matter less,

```{r}
bgb_model_cont_removed_int_removed <- update(bgb_model_cont_removed, .~.-Snail.Level:Nitrogen.level, contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum"))
Anova(bgb_model_cont_removed_int_removed, type="III")
```

but you shouldn't report main effects from the model with the
interaction.

Compare to these to what we actually see (always a good idea!)

#### Plotting outcomes

Results from two-way ANOVAs are often plotted similarly to one-way
ANOVAs, but with colors or other aesthetics representing the additional
group.

```{r}
sdm_summary <- summarySE(valdez_2023[valdez_2023$Snail.Level != "uncaged",], measurevar = "below.biomass.g.meter.sq..m2..", groupvars = c("Snail.Level", "Nitrogen.level")) 
```

```{r}
sdm_summary$Snail.Level <- relevel(sdm_summary$Snail.Level, "removal")  
sdm_summary$Nitrogen.level <- revalue(sdm_summary$Nitrogen.level, c("Fertilized" = "Yes",                                                                     "without"= "No")) 
ggplot(sdm_summary, aes(x=Snail.Level,                            
                        y=below.biomass.g.meter.sq..m2..,  
                        fill=Nitrogen.level)) +   
  geom_col(color="black", position=position_dodge()) +   
  geom_errorbar(aes(ymin=below.biomass.g.meter.sq..m2.., ymax=below.biomass.g.meter.sq..m2..+ci), 
                position = position_dodge()) +   
  labs(title="Grazing impacts depend on nitrogen levels",        
       x= "Grazing level",        
       y= expression(paste("Standing dry mass (" , g^{-1}, m^{-2}, ")")),        
       fill = "Fertilized?")
```

In general, an interaction is not apparent, fertilized plots tend to do
better than unfertilized, and there *might* be less dry mass as snails
increase.

Regardless, we can interpret main effects (though with possibly
different outcomes). The benefit of approach 2 (removing the
interaction) is we "increase" the degrees of freedom associated with the
residuals, which ends up reducing the the MST. This is because in 2-way
ANOVAs we allocate degrees of freedom to calculating main effects and
interactions. This approach was likely used in the original manuscript.

Simply using the the provided output (approach 1) and not performing
another series of tests, however, reduces the chances for a Type 1
error. We will return to this discussion when we get to model selection.

If we see significant effects of a factor that has more than 2 levels
(like we do when using the approach that drops insignificant
interactions), we can consider the general impact of grazing levels
using post-hoc tests:

```{r}
summary(glht(bgb_model_cont_removed_int_removed, linfct = mcp(Snail.Level = "Tukey")))

```

Note the contrast approach has minimal impact here

```{r}
summary(glht(lm(below.biomass.g.meter.sq..m2..~Snail.Level + Nitrogen.level
                , valdez_2023[valdez_2023$Snail.Level != "uncaged",], contrasts = list(Snail.Level = "contr.sum", Nitrogen.level = "contr.sum")), linfct = mcp(Snail.Level = "Tukey")))
```

#### When significant

If the interaction term is significant, it means the main effects can
not be interpreted. This is because the impact of a given variable
depends on the level of another. When this happens, the data is
typically divided into subset and analyzed using one-way ANOVAs.

For example, when @valdez2023 analyzed standing dead mass, they found a
significant interaction term:

```{r}
sdm_model <-lm(Standing.Dead..dry..m2.~Snail.Level * Nitrogen.level, valdez_2023[valdez_2023$Snail.Level != "uncaged",])
plot(sdm_model)
Anova(sdm_model, type="III")
```

Following this, you could investigate impacts in plots with nitrogen,
where you find snail manipulation had a significant impact (and
considered post-hoc which ones!)

```{r}
sdm_model_fertilized <-lm(Standing.Dead..dry..m2.~Snail.Level, valdez_2023[valdez_2023$Snail.Level != "uncaged" & valdez_2023$Nitrogen.level == "Fertilized",])
plot(sdm_model_fertilized)
Anova(sdm_model_fertilized, type= "III")
summary(glht(sdm_model_fertilized, linfct = mcp(Snail.Level= "Tukey")))
```

and plots without added nutrients, where you find snail addition did not

```{r}
sdm_model_not_fertilized <-lm(Standing.Dead..dry..m2.~Snail.Level, valdez_2023[valdez_2023$Snail.Level != "uncaged" & valdez_2023$Nitrogen.level == "without",])
plot(sdm_model_not_fertilized)
Anova(sdm_model_not_fertilized, type= "III")

```

Other approaches for dealing with significant interactions include
directly interpreting interaction terms (which we can do given our
understanding of linear model coefficients), but this is rarely used.
They are somewhat messy

```{r}
coef(sdm_model)
```

Another approach when interactions are significant is to compare all
group means (somewhat like a Tukey design for combined treatment
levels). The **emmeans** package offers this approach.

```{r}
library(emmeans)
emmeans(sdm_model, pairwise ~ Snail.Level*Nitrogen.level)

```

This was likely the approach used in the Valdez et al. 2023 paper.

```{r}
paged_table(data.frame(emmeans(sdm_model, pairwise ~ Snail.Level*Nitrogen.level)$contrasts))
paged_table(data.frame(emmeans(sdm_model, pairwise ~ Snail.Level)$contrasts))
paged_table(data.frame(emmeans(sdm_model, pairwise ~ Nitrogen.level)$contrasts))

```

Note the warning; if interactions are significant comparing main effects
may be inappropriate (which is why other approaches include subsetting
the data).

## Other options

Bootstrapping and permutation tests options may also be used for two-way
ANOVAs when assumptions are not met, though there is implementation is
more complicated than single-sample designs due to the need to
randomize/permute interaction impacts.

## Plotting outcomes

Results from two-way ANOVAs are often plotted similarly to one-way
ANOVAs, but with colors or other aesthetics representing the additional
group.

```{r}
sdm_summary <- summarySE(valdez_2023[valdez_2023$Snail.Level != "uncaged",], measurevar = "Standing.Dead..dry..m2.", groupvars = c("Snail.Level", "Nitrogen.level"))
sdm_summary
```

```{r}
sdm_summary$Snail.Level <- relevel(sdm_summary$Snail.Level, "removal")

sdm_summary$Nitrogen.level <- revalue(sdm_summary$Nitrogen.level, c("Fertilized" = "Yes",
                                                                    "without"= "No"))

ggplot(sdm_summary, aes(x=Snail.Level, 
                           y=Standing.Dead..dry..m2.,
                           fill=Nitrogen.level)) +
  geom_col(color="black", position=position_dodge()) +
  geom_errorbar(aes(ymin=Standing.Dead..dry..m2., ymax=Standing.Dead..dry..m2.+ci), position = position_dodge()) +
  labs(title="Grazing impacts depend on nitrogen levels",
       x= "Grazing level",
       y= expression(paste("Standing dry mass (" , g^{-1}, m^{-2}, ")")),
       fill = "Fertilized?")
```

## Next steps

In the next chapters we will carry our linear model approach to consider
the relationship between continuous outcomes and continuous predictor
variables.
