---
title: One sample tests for continuous data
subtitle: The one where we fully introduce normality
bibliography: ../references.bib
editor: 
  markdown: 
    wrap: 72
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
source("../globals.R")

```

In this chapter will build on our introduction to significance testing
by considering tests for continuous data collected on one trait from a
single population. This will also allow/require us to more fully define
*normal* distributions, which we have already started to discuss.

## Example

Let's return to our iris data and focus on sepal lengths of *I.
viriginica*.

```{r}
set.seed(42)
library(ggplot2)
ggplot(iris[iris$Species == "virginica",],
              aes(x=Sepal.Length)) +
  geom_histogram( fill="blue", color="black") +
  labs(title=expression(paste("Sepal lengths of ",italic("I. virginica"))),
       x= "Sepal length (cm)",
       y= "Frequency")
```

What if we wanted to test if the height was equal to a certain value
like 7 cm?

**We can't, and it's not**.

Height is a random variable. It differs among individuals (see above),
so it isn't equal to any specific value. This may seem obvious, but it's
an important step in understanding hypothesis testing. Many students
also struggle with this, but most are with the fact we learned about
hypothesis testing focusing on proportions (remember the last chapter?).
When we focused on binomial data, it was obvious a single draw could not
be 2 things - it was a success or failure, and we focused on the
relative occurrence of those.

Similarly, for continuous numeric data (remember: data that can on any
value in a given range), we need to focus on describing the distribution
of the data. If we do that, we might want (and be able to test) if, for
example, the *mean* height of *I. virginica* is equal to 7 cm. In fact,
we typically focus on the *mean* of the distribution (one of measures of
central tendency)

To do this, we need to do what we did with binomial data: develop a null
hypotheses, use it to construct a null distribution, and compare our
data to it see how unusual it is (and get a p-value).

<details>

<summary>What is our null hypothesis for this example?</summary>

For this example, we are focused on a two-tailed test (we are asking if
the mean is equal to a certain value), so we have

$$
\begin{split}
H_O: \mu_{sepal \ length} = 7 \ cm \\ 
H_A: \mu_{sepal \ length} \neq 7 \ cm 
\end{split}
$$

</details>

Now that we have a null hypothesis, we need to test it. We can do this
by simulation. Let's make a distribution where the
$\mu_{sepal \ length} =7 \  cm$, then draw samples from it and see how
rare it is to get what we actually observed in the data...which was

```{r}
mean(iris[iris$Species == "virginica","Sepal.Length"])
```

Seems easy enough, but what distribution do we draw from? For our
binomial data we knew exactly what to parameterize - that's because the
entire distribution that fits the null hypothesis is described by the
parameter *p* that is set by the null hypothesis (go back to the last
chapter and note we can find the spread using this one variable as
well!).

If you remember the central limit theorem, you might realize the
distribution of the data does not matter in some ways. No matter what it
looks like, the means of the data will tend towards normality. However,
we still need to describe the data itself it to simulate our draws.
We'll discuss an alternative where you sample from the data itself at
the end of this chapter.

It turns out the shape of the data itself appears fairly normal. So far
we've said normal distributions

-   are roughly symmetric, with tails on either side. Values near the
    middle of the range are more common, with the chance of getting
    smaller or larger values declining at an increasing rate

95% of the sample is within \\\~2 standard deviations of the mean (and
for our mean of means, 95% of the data is within 2 standard errors)

Look at a density function fit to the data:

```{r}
ggplot(iris[iris$Species == "virginica",],
              aes(x=Sepal.Length)) +
  geom_histogram(aes(y = ..density..),fill="blue", color="black") +
  geom_density()+
  labs(title=expression(paste("Sepal lengths of ",italic("I. virginica"))),
       x= "Sepal length (cm)",
       y= "Density")
```

It does appear to be fairly symmetric and peaked in the middle. It turns
out normal distributions (finally defined below!) are very common in
nature.

Just like the binomial data, a normal distribution can be described
using a formula. The formula has 2 parameters that define the shape of
the distribution. $\mu$ defines the center of the distribution, and
$\sigma^2$ describes its spread. The formula for the probabilty density
function is

$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}}
e^{\frac{-(x-\mu)^2}{2\sigma^2}} 
$$

This looks complicated, but remember: this is just an equation for
describing the likelihood of outcomes in a probability space! The first
part $\sqrt{2\pi\sigma^2}$ arises from trying to work with a curve. To
understand the rest, lets take the ln of both sides

$$
\begin{split}
\ln(f(x)) = \ln(\frac{1}{\sqrt{2\pi\sigma^2}}) - \frac{1}{2\sigma^2}*(x-\mu)^2 \\
\ln(f(x)) = -\ln(\sqrt{2\pi\sigma^2}) - \frac{1}{2\sigma^2}*(x-\mu)^2
\end{split}
$$

This may not look like it helps much, but we now have the formula for a
straight line, $y=mx+b$, where

$$
\begin{split}
y= \ln(f(x)) \\
b = - \frac{1}{2\sigma^2}\\
m = -\ln(\sqrt{2\pi\sigma^2})\\
x =  (x-\mu)^2
\end{split}
$$

In other words, our independent variable is the squared distance from
the mean (so all positive)! Note both the y-intercept (the amplitude)
and slope (shape) depends on how spread out the data is ($\sigma^2$).
Note in general when ln(y) decreases linearly with x, y decreases at a
constant proportional rate with x. So we can say a normal random variate
is any random variable in which the probability of an observation
declines in proportion to its squared deviation from the mean (µ).

A normal distribution can take on a lot of shapes given these two
parameters. We can shift where the distribution is centered using the
$\mu$ parameter and how peaked (or not) using the $\sigma$ parameter.

```{r}
#| fig-cap: Note how $\mu$ sets the location of the distribution and $\sigma$ effects the spread
ggplot() + xlim(-20, 20)+
  geom_function(aes(colour = "(0,1)"), fun = dnorm, args = list(mean = 0, sd = 1)) +
  geom_function(aes(colour = "(10,1"), fun = dnorm, args = list(mean = 10, sd = 1)) +
  geom_function(aes(colour = "(0,4)"), fun = dnorm, args = list(mean = 0, sd = 4)) +
  geom_function(aes(colour = "(10,4)"), fun = dnorm, args = list(mean = 10, sd = 4))+
  labs(col = expression(paste('Values of (', mu, ",", sigma, ")" )),
       y="Probability")
```

Let's fit a normal distribution to our data:

```{r}
colors <- c("PDF from data" = "black", "normal curve" = "red")
ggplot(iris[iris$Species == "virginica",],
              aes(x=Sepal.Length)) +
  geom_histogram(aes(y = ..density..),fill="blue", color="black") +
  geom_density(aes(color="PDF from data"))+
    labs(title=expression(paste("Sepal lengths of ",italic("I. virginica"))),
       x= "Sepal length (cm)",
       y= "Density",
       color="Source" ) +
stat_function(fun = dnorm, args = list(mean = mean(iris[iris$Species == "virginica","Sepal.Length"]), sd = sd(iris[iris$Species == "virginica","Sepal.Length"])), aes(color="normal curve"))+
      scale_color_manual(values = colors)

```

Note it fits fairly well, so we'll use it for our sampling experiment.
To do so, we'll take 50 draws (since we had a sample size of 50) from a
normal distribution, find means for each draw, then consider their
distribution.

```{r}
number_of_draws <- 50
number_of_simulations <- 1000

sampling_experiment<- data.frame("observed_mean" = rep(NA, number_of_simulations))
for(i in 1:number_of_simulations){
sampling_experiment$observed_mean[i] = mean(rnorm(50, 7, sd = sd(iris[iris$Species == "virginica","Sepal.Length"])))
}

```

Let's check out the first few outcomes

```{r}
head(sampling_experiment)
```

and plot them

```{r}
ggplot(sampling_experiment,
              aes(x=observed_mean)) +
  geom_histogram(color="black") +
  labs(title="Observed means from 1000 random draws",
       x= "Mean",
       y= "Frequency")
```

Now let's see how that compares to what we actually saw.

```{r}
sampling_experiment$compare =   ifelse(abs(sampling_experiment$observed_mean-7) >= abs(mean(iris[iris$Species == "virginica","Sepal.Length"])-7), 'as or more extreme', 'not as or more extreme')

sampling_experiment$compare <- factor(sampling_experiment$compare)
levels(sampling_experiment$compare) <- c(levels(sampling_experiment$compare), "as or more extreme")

ggplot(sampling_experiment,
              aes(x=observed_mean, fill=compare)) +
  geom_histogram(color="black") +
  labs(title="Observed means from 1000 random draws",
       x= "Mean",
       y= "Frequency", 
       fill = "Sampled mean is ...") +
    scale_fill_discrete(drop = FALSE)

```

So in our example simulation, **no** observed means were as far from the
value from the null hypothesis as our sample mean was. This would leave
to a p-value of 0 - unusual in some regards, but possible here.

## Another way to compare

Just like we saw for binomial data, we can always carry out a sampling
experiment to find a p-value. However, that's mainly because we have
computers. Even with computers, it would be cumbersome to set up a new
sampling experiment for every dataset (which we would need to do for any
change in sample size, mean, or standard deviation).

Instead it would be nice to find a way to replicate the distribution we
see above for similar experiments using an equation (just like we did
for the binomial data!). One step in doing this relates to how we
examine our predictions given our hypotheses. It turns out, we can map
our hypotheses to *models* that explain the variation we see in the
data. Our hypotheses (remember

$$
\begin{split}
H_O: \mu_{sepal \ length} = 7 \ cm \\ 
H_A: \mu_{sepal \ length} \neq 7 \ cm 
\end{split}
$$

focus on the mean. They are also stating a prediction for every outcome!
Under the null hypothesis, if we were asked to guess the length of a
sepal from *I. virginica*, we would guess 7 cm. Under the alternative
hypothesis, we would guess it's something different than 7. Using our
sample, we might instead guess it's equal to 7 + $\delta$, where
$\delta$ is estimated from our sample and 7+ $\delta$ is the mean of our
sample.

For each hypothesis, we could calculate a measure of related mode fit
called the sum squared error from our model, or SSE, where

$$
SSE=sum \ squared \ error = \sum_{i=1}^{n}(Y_i-\hat{Y_i})^2
$$

Here, $Y_i$ are the data points, and $\hat{Y_i}$ is our predicted value.

This is basically saying we compare the predicted to observed value for
each data point for each model (the square exists so things don't cancel
out!). Our null hypothesis corresponds to a simpler view of the world (a
reduced or null model), where $\hat{Y_i}$ is equal to a given value (in
our case, 7 cm), whereas under the alternative hypothesis (which
corresponds to an alternative or full model), $\hat{Y_i}$ is equal to a
different value (such as 7 + $\delta$). The value for $\delta$ is
estimated from our sample and makes the full model larger than the
reduced in regard to the number of parameters included in the model.

We can then compare the SSE of the 2 models by finding their difference.
This is our **signal** from the data. If we take multiple samples from a
known population that is defined by the null hypothesis, we can carry
out a very similar sampling to what we did originally.

```{r}

for(i in 1:number_of_simulations){
a <- rnorm(50, 7, sd = sd(iris[iris$Species == "virginica","Sepal.Length"]))
sampling_experiment$observed_mean[i] <- mean(a)

sampling_experiment$SSE_null[i] <-  sum((a-7)^2)

sampling_experiment$SSE_full[i] <- sum((a-mean(a))^2)
}

sampling_experiment$SSE_difference <- sampling_experiment$SSE_null - sampling_experiment$SSE_full

ggplot(sampling_experiment,
              aes(x=SSE_difference)) +
  geom_histogram(color="black") +
  labs(title="Observed difference in model fit from 1000 draws",
       x= "Mean",
       y= "Frequency")


difference_SSE_observed <- sum((iris[iris$Species == "virginica","Sepal.Length"]-7)^2)-sum((iris[iris$Species == "virginica","Sepal.Length"]-mean(iris[iris$Species == "virginica","Sepal.Length"]))^2)

ggplot(sampling_experiment,
              aes(x=SSE_difference)) +
  geom_histogram(color="black") +
  labs(title="Observed difference in model fit from 1000 draws",
       x= "Mean",
       y= "Frequency")+
  geom_vline(aes(xintercept=difference_SSE_observed))

```

Note we never saw as large a difference in the **signal** our sampling
experiment as we did in the data!

```{r}
nrow(sampling_experiment[sampling_experiment$SSE_difference >=  difference_SSE_observed,])
```

Note this distribution is skewed for a few reasons. Given the square
term, it must positive. We also unlikely (given sampling error) that
signal will be exactly zero. In fact,since we estimate $\delta$ from the
data itself, the alternative model is almost always a better fit
(remember our bias in variance estimates?).

This approach accounts for the **noise** that we see in our data
(variation we expect to see in signal values even if the null hypothesis
is true) through sampling. However, if we could estimate the noise, we
could also divide our signal by it. Accounting for signal and noise
let's us take very different questions in terms of data and use a
standardized approach to analyze them. This is because a signal to noise
ratio typically follows some distribution.

## Z-test: a distribution based approach

In our case, the signal to noise ratio is approximated by the normal
distribution! This is an *approximate* solution to the signal to noise
ratio because our means approach a normal distribution as sample size
increases - so for non-infinity sample sizes, they may not be perfectly
normal!

We can show that our signal is simply the mean from our data minus the
mean under the null hypothesis, and including $\sigma$ in the
denominator accounts for noise. We can make this more generalizable if
we z-transform the data using the formula

$$
z=\frac{Y - \mu}{\sigma}
$$

After this transformation, the data is centered at 0 (think about it -
if you subtract the mean from all data points...) and has a standard
deviation of 1 (because you divided by the standard deviation!). This
also makes the mean of transformed data equal to $\delta$ and the mean
under the null hypothesis equal to zero. This also means \~68% of the
data points lie between -1 and 1, while \~95% lie between -2 and 2
(since the standard deviation is 1!).

```{r, echo = F}
#| label: fig-z-colored
#| fig-cap: "~65% of the data lies with 1 standard deviation of the mean, ~95% lies within 2 standard deviations, and ~99% lies within 3 standard deviations of the mean"
library(viridis)
colors <- setNames( viridis(5), c("0-1","1-2", "2-3", "3-4", "4+"))


#standard normal with 1-4 se highlighted ####
ggplot(data = data.frame(x = c(-5, 5)), aes(x)) +
   stat_function(aes(fill="4+"), fun = dnorm, args = list(mean = 0, sd = 1),
                geom = "area") +
  stat_function(aes(fill="3-4"),fun = dnorm, geom = "area", xlim =c(-4,4)) +
  stat_function(aes(fill="2-3"),fun = dnorm, geom = "area", xlim =c(-3,3)) +
  stat_function(aes(fill="1-2"),fun = dnorm, geom = "area", xlim =c(-2,2)) +
  stat_function(aes(fill="0-1"),fun = dnorm, geom = "area", xlim =c(-1,1)) +
  scale_x_continuous(breaks=seq(-5,5,1)) +
  labs(y="Probablity",
       x="Z score",
       fill="SE from mean") 
```

We call this specific form of the normal distribution (N(0,1), showing
the mean and standard deviation parameters) the Z distribution.
Extending on this with some algebra, we can get a Z-score from any data
using the equation

$$
z=\frac{\bar{x} - \mu}{\sigma}
$$

This transforms a given data point into a z-score on the z, or standard
normal, distribution. We can then use the Z distribution to consider how
unusual our z-scores are (i.e., get a p-value!). We call the approach
that uses this distribution the z-test, and it can be carried out using
the z.test function in the BDSA package.

```{r}
library(BSDA)
z.test(iris[iris$Species == "virginica","Sepal.Length"], mu = 7, 
             sigma.x= sd(iris[iris$Species == "virginica","Sepal.Length"]))
```

Using this approach we get a p-value of .000005 - not 0, but very close!

Here's another example of when to use a z-test:

The average time taken by bartenders at a local bar to make a party of
20 people drinks is believed to be 30 minutes. The manager at the bar
has been getting complaints about the wait time from the customers and
wants to improve. A sample of 25 employees who completed making 20
drinks after a training program showed an average completion time of 25
minutes with a standard deviation of 2 minutes.

Is there enough evidence to suggest that the new training program has a
significant effect on the average completion time?

```{r}
library(BSDA)
drink_times <- c(28, 34, 25, 40, 22, 13, 30, 31, 29, 23, 15, 39, 44, 37, 28, 24, 35, 32, 26, 17, 17, 20, 33, 27)
z_test <- z.test(x = drink_times, mu = 30, sigma.x = 2, alternative = "two.sided")
print(z_test)

```

### A little history

Although software now provides us exact (approximate) p-values,
historically this was far more difficult. For this reason, people took
advantage of a transformation so they could use a standardized table
like this one.

[![Example of z-table. Jsmura, Creative Commons Attribution-Share Alike
4.0 International
licence](/images/Tabla_z.png){fig-align="center"}](https://commons.wikimedia.org/wiki/File:Tabla_z.png)

These tables showed how Z-scores related to p-values. Note these often
showed the area to the left of the value, so two-tailed tests required
one to multiple the given p-value by 2 (or, if focused on the upper
tail, multiply (1-given p-value) by 2 since the distribution is
symmetric. Similarly, some tables only had values \<0; for those you
could find the score whose absolute value corresponed to the observed
z-score and multiply the noted p-value by 2 for two-tailed tests .

For our example, we got a z score of -4.8515. The table doesn't even go
that low!

## Does the distribution of the data matter?

Remember we are focusing on the distribution of the means (both our
sampling experiment and SSE calculations include the means of the sample
and data under the null hyothesis!). Given that and the central limit
theorem, does the distribution of the data matter? Yes, but only in
regards to the relationship between sample size and normality of the
sample means. If the underlying data is normal, then the sampled means
are distributed normally for almost *any* sample size, although sample
size impacts the spread of the sample means.

```{r, echo = F}
#| label: fig-normal_draws
#| fig-cap:  "Means drawn from a normal distribution are normal regardless of sample size"
number_of_simulations <- 1000
sample_size=c("1","5","10", "20", "40", "80")

sampling_experiment <- setNames(data.frame(matrix(ncol = length(sample_size), nrow = number_of_simulations)), sample_size)
for(k in 1:length(sample_size)){
for(i in 1:number_of_simulations){
sampling_experiment[i,k] = mean(rnorm(n=as.numeric(sample_size[k])))
}
}
library(reshape2)
sampling_experiment_long <- melt(sampling_experiment, variable.name = "Sample_size", value.name = "mean")
sampling_experiment_long$Sample_size <- factor(sampling_experiment_long$Sample_size, levels =c("1","5","10", "20", "40", "80"))
levels(sampling_experiment_long$Sample_size) <- paste("Sample size of ", levels(sampling_experiment_long$Sample_size))

ggplot(sampling_experiment_long,aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title=paste("Observed means from ", number_of_simulations, " random draws"),
       subtitle = "Normal distribution",  
       x= "Mean",
       y= "Frequency")+
    facet_wrap(~Sample_size, nrow = 2)
```

For other distributions, larger sample sizes are required to approximate
normality. For example, consider a highly-peaked (double-exponential)
distribution

```{r, echo=F}
sampling_experiment <- setNames(data.frame(matrix(ncol = length(sample_size), nrow = number_of_simulations)), sample_size)
library(VGAM)
for(k in 1:length(sample_size)){
for(i in 1:number_of_simulations){
sampling_experiment[i,k] = mean(rlaplace(n=as.numeric(sample_size[k])))
}
}

sampling_experiment_long <- melt(sampling_experiment, variable.name = "Sample_size", value.name = "mean")
sampling_experiment_long$Sample_size <- factor(sampling_experiment_long$Sample_size, levels =c("1","5","10", "20", "40", "80"))
levels(sampling_experiment_long$Sample_size) <- paste("Sample size of ", levels(sampling_experiment_long$Sample_size))

ggplot(sampling_experiment_long,aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title=paste("Observed means from ", number_of_simulations, " random draws"),
       subtitle = "Double-exponential distribution",  
       x= "Mean",
       y= "Frequency")+
    facet_wrap(~Sample_size, nrow = 2)
```

or as skewed $\chi^2$ distribution (here with a df =4, to be explained
later!):

```{r, echo = F}

sampling_experiment <- setNames(data.frame(matrix(ncol = length(sample_size), nrow = number_of_simulations)), sample_size)
library(VGAM)
for(k in 1:length(sample_size)){
for(i in 1:number_of_simulations){
sampling_experiment[i,k] = mean(rchisq(n=as.numeric(sample_size[k]), df = 4))
}
}

sampling_experiment_long <- melt(sampling_experiment, variable.name = "Sample_size", value.name = "mean")
sampling_experiment_long$Sample_size <- factor(sampling_experiment_long$Sample_size, levels =c("1","5","10", "20", "40", "80"))
levels(sampling_experiment_long$Sample_size) <- paste("Sample size of ", levels(sampling_experiment_long$Sample_size))

ggplot(sampling_experiment_long,aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title=paste("Observed means from ", number_of_simulations, " random draws"),
       subtitle = "Chi-squared distribution, df = 4",  
       x= "Mean",
       y= "Frequency")+
    facet_wrap(~Sample_size, nrow = 2)
```

Even for these very non-normal distributions, the means approach a
normal distribution at fairly low sample sizes. This is even true for
binomial data, especially when *p* is not very close to 0 or 1. Consider

```{r, echo=F}
sampling_experiment <- setNames(data.frame(matrix(ncol = length(sample_size), nrow = number_of_simulations)), sample_size)

for(k in 1:length(sample_size)){
for(i in 1:number_of_simulations){
sampling_experiment[i,k] = rbinom(n=1, size=as.numeric(sample_size[k]), prob=.7)
}
}

sampling_experiment_long <- melt(sampling_experiment, variable.name = "Sample_size", value.name = "mean")
sampling_experiment_long$Sample_size <- factor(sampling_experiment_long$Sample_size, levels =c("1","5","10", "20", "40", "80"))
levels(sampling_experiment_long$Sample_size) <- paste("Sample size of ", levels(sampling_experiment_long$Sample_size))

ggplot(sampling_experiment_long,aes(x=mean)) +
  geom_histogram(color="black") +
    labs(title=paste("Observed number of successes from ", number_of_simulations, " random draws"),
       subtitle = "Binomial distribution, p=.7",  
       x= "Mean",
       y= "Frequency")+
    facet_wrap(~Sample_size, nrow = 2)
```

This is why it used to be more common to use a normal approximation to
the binomial distribution - even though the binomial distribution is
easy to compute,the z is even easier! For example, Sandidge
[@sandidge2003] noted that brown recluse spiders chose dead prey items
(as opposed to live - 2 categories!) when offered choices. This data
could be assessed using a binomial test

```{r}
 binom.test(119,(59+41+41), p=.5)
 
```

or a z-test approach by finding a z-sore

```{r}
 p=.5
 n=41+41+59
 z_score <- (119-p*n)/sqrt(n*p*(1-p))
 pnorm(z_score, lower.tail = F) *2

```

or a $\chi^2$ test, which is similar to a z test but focuses on a sum of
independent squared Z variates. The $\chi^2$ distribution is defined by
a degrees of freedom, or df, parameter, which in this case isequal to
the number of categories -1 (or 2-1=1 in this case). We will return to
df multiple times! Note the p-values obtained from the Z and $\chi^2$
tests are the same!

```{r}
chisq.test(c(119,n-119))
```

This may seem esoteric, but understanding these issues may help you
interpret older papers while also choosing to employ more modern
statistical methods.

### QQ norm plots - commonly used, but not needed, at this point!

In addition to considering the sample size and underlying distribution,
quantile-quantile (Q-Q) plots are sometimes used to assess normality.
These plots plot quantiles in one data set against quantiles from
another to determine if they come from similar distribution. Remember,
quantiles just order data; percentiles are example where you have 100
cut points. Q-Q norm plots consider if a given dataset are similar to a
normal distribution. If so, then the dots should match the straight line
produced by the *qqline* function.

```{r}
qqnorm(iris[iris$Species == "virginica","Sepal.Length"])
qqline(iris[iris$Species == "virginica","Sepal.Length"])
```

While we introduce Q-Q plots here, and they are often used to assess
normality, remember our tests (so far) are relying on the means of the
data being normally-distributed and not the data itself!

## What if we don't know the variance?

Assuming everything above makes sense, we are left with one issue: the
variance of the underlying population is rarely known!

In the above examples, we actually used our *estimate* of variance from
the sample to run our simulation experiment and z-test! While this works
ok for large sample sizes (yay for central limit theorem!) and is what
statisticians relied upon historically, it doesn't work for well for
smaller sample sizes (unless we somehow know the population variance).
Our estimates for the population variance are less precise and
potentially biased at small sample sizes.

To address this issue, statisticians developed the t-distribution.
Unlike the normal distribution, its shape depends on the sample size.
This parameter is coded as degrees of freedom, commonly denoted as df,
and is equal to n - 1 (we'll come back to df later!). The major
breakthrough, however, was that df was the *only* sample-specific
parameter. The same distribution works regardless of the estimated
population variance, as a t statistic/score is created that functions
like a z score.

$$
t=\frac{\bar{x} - \mu}{\frac{s}{\sqrt{n}}} 
$$

It can be shown (though not here!) that the t-distribution is actually a
specific form of the F distribution (which we'll see in ANOVAs). An
F-distribution is the ration of two $\chi^2$ distributions, which (as
noted above) are sums of squared Z distributions. The t distribution is
the special case where you can take the square of an F distribution
where the numerator (top) $\chi^2$ distribution has 1 degree of freedom
and the denomintor (bottom) has n-1 degrees of freedom. In general,
variance follows a $\chi^2$ distribution with n-1 degrees of freedom.

Because it directly uses the estimate of the population variance,
smaller sample sizes show more spread (and thus make null hypotheses
more difficult to reject!). For example, note how small sample sizes
(remember, df=3 means n=4!) are notably different from the normal
distribution, while larger sample sizes become very hard to distinguish!

```{r, echo=F}
colors <- setNames( viridis(5), c("normal","df=3", "df=10", "df=20", "df=50"))
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(aes(color="normal"),fun = dnorm, args = list(mean = 0, sd = 1), 
                geom = "line") + 
  stat_function(aes(color="df=3"),fun = dt, args = list( df= 3), 
                geom = "line") +
  stat_function(aes(color="df=10"),fun = dt, args = list( df= 10), 
                geom = "line") + 
  stat_function(aes(color="df=20"),fun = dt, args = list( df= 20),
                geom = "line") +
  stat_function(aes(color="df=50"),fun = dt, args = list( df= 50), 
                geom = "line") +
  labs(y="Probablity",
       x="Score (t or z)",
       color="Distribution")
```

Note this means the t-distribution is actually a class of distributions.

In older text books, you would have a t-table that showed t scores
corresponding to commonly used values of $\alpha$ for multiple degrees
of freedom. These published t scores are sometimes called critical
values. Users would have compared their calculated t-score (or its
absolute value in the case of negative values) to the appropriate
critical values (often $\alpha/2$ for 2-sided tests) to determine if a
finding was significant. We can actually produce a t table in R.

```{r}
#| label: fig-t_table
#| fig-cap: " Critical values"
df <- 1:100
cut_offs <- c(".1", ".05", ".025", ".01")
t_table <- setNames(data.frame(matrix(ncol = length(cut_offs)+1, nrow = length(df))), c("df",cut_offs))
t_table$df <- df
for (i in 1:length(df)){
  for (j in 2:ncol(t_table))
  t_table[i,j] <- round(abs(qt(as.numeric(colnames(t_table)[j]), df[i])),3)
}
library(rmarkdown)
paged_table(t_table)
```

To calculate the test statistic in R, we can instead (thankfully!) use
the *t.test* function.

```{r}
t.test(iris[iris$Species == "virginica","Sepal.Length"], mu = 7)
```

Note the t-test also provides confidence intervals. Note these consider
the spread of the data given the t-distribution so they are always wider
than those predicted using a normal distribution, though the difference
is small at large sample sizes.

This explains why I have noted \~95% of the data lies within 2 standard
errors of the mean! For truly normal data, it's actually within 1.96
standard errors. For data where we estimate the population variance, it
depend on the sample size, but even at n=21 (and thus df = 20!) the
number is 2.09. Note

```{r}
t_critical <- setNames(data.frame(matrix(ncol = 2, nrow = length(sample_size))), c("n", "95% of the data is within this many standard deviations of the mean"))
for (i in 1:length(sample_size)){
t_critical$n[i] <- sample_size[i]
t_critical[i,2] <-qt(.025,as.numeric(sample_size[i]))
}

```

## What if we don't trust the normal approximation?

Despite the central limit theorem (or because of it), we may not think
our sample size is sufficient given the distribution the data to assume
that $\hat{Y}$ really follows a normal distribution. In that case we a
few options.

### Wilcoxon test (aka, signed binary-transform, Fisher's sign test)

If the distribution of the data is symmetric, a Wilcoxon test may be
appropriate. **Note it is rare to have data that are symmetrically
distributed but that for which you don't think th means will be normally
distributed)**. We introduce the test here as it will come back up (and
be more useful) in later chapters.

This test employs a strategy we will see again and again: it ranks the
data, in this case based on the distance from the mean under the null
hypothesis. A sign is also assigned to each rank, with those originating
from data points that were lower than the proposed mean becoming
negative. In theory, the sum of the signed ranks should be \~0 if H~O~
is true. We can carry out this test in R using the *wilcox.test*
function.

```{r}
wilcox.test(iris[iris$Species == "virginica","Sepal.Length"], mu = 7)
```

Here's another example of when to use a Wilcoxon test:

A hair stylist is studying the growth of hair after use of her new
stimulating oil. The stylist believes that the average growth of her
clients should be 2 inches after a month. To test this hypothesis, the
stylist takes a sample of 15 women and measured their hair length. The
collected data is as follows:

\[2.5,3.2,1.8,1.1,0.5,3.3,1.9,3.0,0.4,2.2,1.7,2.3,1.1,3.8,4.0\]

Is there enough evidence to suggest that the average growth of hair is
significantly different from 2 inches after using the stimulating oil?

```{r}
hair_length <- c(2.5,3.2,1.8,1.1,0.5,3.3,1.9,3.0,0.4,2.2,1.7,2.3,1.1,3.8,4.0)
wilcoxon_test_result <- wilcox.test(hair_length, mu = 2)
print(wilcoxon_test_result)
```

### Sign test (aka the median test)

If the data are not symmetrically distributed, the sign test actually
just counts those below the proposed value (which is the median here
instead of the mean, since we are concerned about normality). In theory,
approximately half the values should be under the proposed mean if H~O~
is true. The proportion below is compared to .5 using a binomial test.

```{r}
SIGN.test(iris[iris$Species == "virginica","Sepal.Length"], md = 7)
```

Here's another example of how to use the sign test:

The median income of high school teachers in two different cities, the
Bronx and Brooklyn, is being compared. The board of education believes
that the median income in the Bronx A is \$50,000. To investigate this
claim, a random sample of 25 teachers from the Bronx and 25 residents
from Brooklyn is taken, and their incomes are recorded.

In the Bronx, the median income is \$45,000, and the data is as follows:
\[42, 48, 50, 38, 40, 43, 49, 52, 47, 41, 44, 46, 45, 49, 55, 43, 46,
48, 39, 50\]

In Brooklyn, the median income is \$55,000, and the data is as follows:
\[60,58,56,53,55,62,57,58,54,55,56,59,61,57,53,56,60,58,59,62,55,57,58,59,60\]

Is there enough evidence to suggest that the median income in either
neighborhood A or B is significantly different from the board of
education's claim?

```{r}
library(BSDA)
bronx_income <- c(42, 48, 50, 38, 40, 43, 49, 52, 47, 41, 44, 46, 45, 49, 55, 43, 46, 48, 39, 50, 34, 65, 70, 45, 50)
brooklyn_income <- c(60, 58, 56, 53, 55, 62, 57, 58, 54, 55, 56, 59, 61, 57, 53, 56, 60, 58, 59, 62, 55, 57, 58, 59, 60)
median_test_result <- SIGN.test(x = bronx_income, y = brooklyn_income, md = 50000)
print(median_test_result)
```

### Bootstrapping

The final option we will review is a little different. For all our other
hypothesis tests we've been resampling from a distribution that fits the
parameters from the null hypothesis. However, its turns out we can
resample from the actual data we collected to approximate the
distribution of the sample (or the signal in most cases). You can then
use that distribution to develop confidence intervals or hypothesis
testing. The only requirement here is we have a large enough sample size
to actually appropriately sample from (and that we have the means to do
it!). This approach was developed in the 1990s given the increase in
computing power and availability.

We can demonstrate this with an imaginary population (and also
demonstrate the central limit theorem). Let's make a population whose
trait value of focus falls between 60 and 80 in a uniform manner.

```{r}
population_unif <- data.frame(id = 1:10000, 
                              value = runif(10000, 60, 80))
ggplot(population_unif, aes(x=value)) +
  geom_histogram(color="black") +
  labs(title="Our imaginary population",
       x= "Trait value",
       y= "Frequency")

```

Now's let's take samples of 50 from it a lot of times
(`r number_of_simulations` here) and plot the means of these samples.

```{r}
sampling_experiment<- data.frame("observed_mean" = rep(NA, number_of_simulations))
for(i in 1:number_of_simulations){
sampling_experiment$observed_mean[i] <- mean(sample(population_unif$value,50, replace = F))
}
ggplot(sampling_experiment, aes(x=observed_mean)) +
  geom_histogram(color="black") +
  labs(title="Means from imaginary samples of size 50 from our imaginary population",
       x= "Trait value",
       y= "Frequency")
```

Now let's instead think of what might happen - we get a single sample of
50.

```{r}
actual_sample <- data.frame(sample = sample(population_unif$value,50, replace = F))
ggplot(actual_sample, aes(x=sample)) +
  geom_histogram(color="black") +
  labs(title="Actual sample of size 50 from our imaginary population",
       x= "Trait value",
       y= "Frequency")
```

Now (here's the odd part): Let's sample from that sample with
replacement to get "new" samples, then get those means, again,
`r number_of_simulations` times.

```{r}
bootstrap_outcomes <- data.frame(mean = rep(NA, number_of_simulations), sd = NA)
for (i in 1:number_of_simulations){
example_bootstrap <-sample(actual_sample$sample, length(actual_sample$sample), replace = T)
bootstrap_outcomes$mean[i] <- mean(example_bootstrap)
bootstrap_outcomes$sd[i] <- sd(example_bootstrap)
}
```

When we plot them we see it looks very much like the distribution of
means we obtained by re-sampling!

```{r}
ggplot(bootstrap_outcomes, aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title="Bootstrapped means of size 50 from our imaginary population",
       x= "Trait value",
       y= "Frequency")
```

Overall, this means we can use our sample to recreate our underlying
distribution and explore its properties!

Let's demonstrate this approach with our iris data. First, we can make
"new" datasets from our original data by sampling (with replacement)
samples of the same size from our original dataset.

```{r}
bootstrap_data<- iris[iris$Species == "virginica","Sepal.Length"]
bootstrap_outcomes <- data.frame(mean = rep(NA, number_of_simulations), sd = NA)
for (i in 1:number_of_simulations){
iris_bootstrap <-sample(bootstrap_data, length(bootstrap_data), replace = T)
bootstrap_outcomes$mean[i] <- mean(iris_bootstrap)
bootstrap_outcomes$sd[i] <- sd(iris_bootstrap)
}
ggplot(bootstrap_outcomes, aes(x=mean)) +
  geom_histogram(color="black") +
  labs(title=expression(paste("Bootstrapped means of sepal lengths of ",italic("I. virginica"))),
       x= "Mean sepal length (cm)",
       y= "Frequency")

```

We again (thanks to the central limit theorem) see the means follow a
normal distribution.

We can also carry this out using the *boot* function (in the **boot**
package), though the functions may look a little odd.

```{r}
library(boot)
results <- boot(data=bootstrap_data, statistic = function(x, inds) mean(x[inds]),
   R=number_of_simulations)
ggplot(data.frame(results$t), aes(x=results.t)) +
  geom_histogram(color="black") +
  labs(title=expression(paste("Bootstrapped means of sepal lengths of ",italic("I. virginica"))),
       x= "Mean sepal length (cm)",
       y= "Frequency")
```

and then find the 95% confidence interval. Like binomial data, there are
a few ways to do this. One is to use the percentile, or quantile,
method. This is intuitive. We rank the bootstrapped values from smallest
to largest and then find points that cut off the bottom and top 2.5%.

```{r}
quantile( results$t, probs=c(.025, .975) ) 

```

Though simple, these findings may also be biased. More advanced
intervals are provided the *boot.ci* function.

```{r}
boot.ci(results)

```

The *boot.t.test* function in the **MKinfer** package offers another way
to calculate bootstrap statistics for single-sample continuous data. It
returns the percentile confidence intervals and also offers a p value.

```{r}
library(MKinfer)
boot.t.test(iris[iris$Species == "virginica","Sepal.Length"], mu = 7)
```

## Next steps

Now that we've covered dealing with categorical and continuous data, we
will move to comparing populations to each other.
